---
format: gfm
---

# footbet

European Football Betting Analytics with Sharp Odds.

Download, process, and model European football match data and sharp bookmaker
odds (Pinnacle). Implements Poisson/Dixon-Coles models, Elo ratings, and
Kelly-criterion staking for value betting on full-time and half-time outcomes.

## Installation

### Standard R

```r
# install.packages("remotes")
remotes::install_github("johngavin/footbet")
```

### Nix (Recommended)

```bash
git clone https://github.com/johngavin/footbet.git
cd footbet
./default.sh
```

### rix Integration

```r
library(rix)
rix(
  r_pkgs = c("dplyr", "arrow", "duckdb"),
  git_pkgs = list(
    list(
      package_name = "footbet",
      repo_url = "https://github.com/johngavin/footbet",
      commit = "HEAD"
    )
  ),
  ide = "none",
  project_path = ".",
  overwrite = TRUE
)
```

## Coverage

10 leagues (top 2 divisions) across 5 countries, 10 years of history:

| Country | Division 1 | Division 2 |
|---------|-----------|-----------|
| England | E0 (Premier League) | E1 (Championship) |
| Germany | D1 (Bundesliga) | D2 (2. Bundesliga) |
| Italy | I1 (Serie A) | I2 (Serie B) |
| Spain | SP1 (La Liga) | SP2 (Segunda) |
| France | F1 (Ligue 1) | F2 (Ligue 2) |

## Model Pipeline

```{r}
#| echo: false
#| output: asis
if (file.exists(here::here("_targets"))) {
  targets::tar_config_set(store = here::here("_targets"))
  tryCatch({
    targets::tar_load(readme_pipeline_mermaid)
    cat(readme_pipeline_mermaid)
  }, error = function(e) {
    # Fallback to text description when target not built
    cat("1. **Data**: football-data.co.uk match results + Pinnacle closing odds\n")
    cat("2. **Features**: Rolling goals, Elo ratings, devigged implied probabilities\n")
    cat("3. **Models**: Poisson GLM baseline, Dixon-Coles, Elo-based\n")
    cat("4. **Evaluation**: Walk-forward CV with log-loss, Brier score, calibration\n")
    cat("5. **Decisions**: Fractional Kelly staking with drawdown guardrails\n")
  })
} else {
  cat("1. **Data**: football-data.co.uk match results + Pinnacle closing odds\n")
  cat("2. **Features**: Rolling goals, Elo ratings, devigged implied probabilities\n")
  cat("3. **Models**: Poisson GLM baseline, Dixon-Coles, Elo-based\n")
  cat("4. **Evaluation**: Walk-forward CV with log-loss, Brier score, calibration\n")
  cat("5. **Decisions**: Fractional Kelly staking with drawdown guardrails\n")
}
```

## Walk-Forward AH Backtest: Null Result

A full walk-forward Asian Handicap backtest against Pinnacle prices across
six seasons and four models (GLM, Ranger, xG-Kalman-Dixon-Coles, ensemble),
with both flat and Kelly staking, did not find a profitable edge. Best ROI
reached -5.4% after Kelly staking, per-league intercepts and Kalman σ tuning
(see issues [#59](https://github.com/JohnGavin/footbet/issues/59),
[#81](https://github.com/JohnGavin/footbet/issues/81),
[#82](https://github.com/JohnGavin/footbet/issues/82)). The xG-Kalman-DC
stack was the second-best model at -7.3% before tuning.

- **Detailed methodology, results and leaderboard**: [`plans/MODEL_CATALOGUE.md`](plans/MODEL_CATALOGUE.md)
- **Session-by-session lab notes and failed approaches**: [`CHANGELOG.md`](CHANGELOG.md)
- **Closing Line Value (CLV) diagnostic**: `R/clv.R` — `load_closing_ah_prices()`,
  `attach_clv()`, `summarise_clv()`. See `plan_clv.R` for targets
  `oos_ah_clv_summary`, `oos_ah_{ranger,ensemble}_clv_summary`. CLV is the
  honest follow-up test to "no edge vs Pinnacle" — it disambiguates
  "model found nothing" from "model found something only softer books pay".
- **Kalman innovation logging**: `kalman_strengths(..., record_innovations = TRUE)`
  returns standardised innovations `ε_t = (y_t - ŷ_t|t-1) / sqrt(S_t)` for
  regime-break diagnostics (manager changes, transfer shocks).

Publishing a documented null result from a clean walk-forward backtest is
rare in this space; the intent is that other projects attempting the same
stack can calibrate expectations against these results rather than
rediscovering them.

## Positive Finding: xG Features Improve Calibration

While the AH bet family produces no edge, xG rolling features do
reliably improve prediction quality over a goals-only Poisson GLM
baseline in leak-corrected walk-forward CV
([#84](https://github.com/JohnGavin/footbet/issues/84)):

- **Log-loss**: goals-only 1.07 → xG 1.00 (~6.5% improvement)
- **Brier**: 0.620 → 0.592 (~4.5%)
- **RPS**: 0.211 → 0.202 (~4.3%)

The improvement survives the 7-day bet-time cutoff correction
unchanged (cut0 vs cut7: log_loss 1.00 vs 1.00).

**Per-league breakdown** (log-loss improvement, cut7):
EPL +1.6%, Serie A +1.6%, Bundesliga +1.3%, Ligue 1 +0.9%, La Liga +0.2%.
All 5 leagues show improvement; none regress.

**Model comparison on xG-era data** (top-5 leagues, 2019–2025):
GLM goals-only 1.018, Dixon-Coles 1.017, GLM+xG 1.003.
Dixon-Coles barely beats GLM on goals alone; xG features provide a
larger gain than the DC low-score correction.

**Brier decomposition**: xG improves resolution (0.019 vs 0.019) while
maintaining identical reliability (~0.0006). The gain is in sharpness
(more decisive predictions), not better calibration. The xG advantage
is real but too diffuse to overcome Pinnacle's AH vig when translated
to per-match bet selection.

## Project Structure

```{r}
#| echo: false
#| output: asis
# Filter via dir_ls()'s regexp/invert (fs::dir_tree forwards ...), NOT by
# grepping the rendered lines: dir_tree prints children as bare indented
# basenames, so a line filter drops the "_targets" header while leaving all
# ~250 of its descendants ("│   ├── objects", "│   │   ├── ah_roi_per_team").
tree <- capture.output(
  fs::dir_tree(recurse = 2, regexp = "_targets", invert = TRUE)
)
cat("<details>\n<summary>Project structure (click to expand)</summary>\n\n```\n")
cat(tree, sep = "\n")
cat("\n```\n\n</details>\n")
```

## License

MIT
