---
format: gfm
---

# footbet

European Football Betting Analytics with Sharp Odds.

Download, process, and model European football match data and sharp bookmaker
odds (Pinnacle). Implements Poisson/Dixon-Coles models, Elo ratings, and
Kelly-criterion staking for value betting on full-time and half-time outcomes.

## Installation

### Standard R

```r
# install.packages("remotes")
remotes::install_github("johngavin/footbet")
```

### Nix (Recommended)

```bash
git clone https://github.com/johngavin/footbet.git
cd footbet
./default.sh
```

### rix Integration

```r
library(rix)
rix(
  r_pkgs = c("dplyr", "arrow", "duckdb"),
  git_pkgs = list(
    list(
      package_name = "footbet",
      repo_url = "https://github.com/johngavin/footbet",
      commit = "HEAD"
    )
  ),
  ide = "none",
  project_path = ".",
  overwrite = TRUE
)
```

## Coverage

10 leagues (top 2 divisions) across 5 countries, 10 years of history:

| Country | Division 1 | Division 2 |
|---------|-----------|-----------|
| England | E0 (Premier League) | E1 (Championship) |
| Germany | D1 (Bundesliga) | D2 (2. Bundesliga) |
| Italy | I1 (Serie A) | I2 (Serie B) |
| Spain | SP1 (La Liga) | SP2 (Segunda) |
| France | F1 (Ligue 1) | F2 (Ligue 2) |

## Model Pipeline

```{r}
#| echo: false
#| output: asis
if (file.exists(here::here("_targets"))) {
  targets::tar_config_set(store = here::here("_targets"))
  tryCatch({
    targets::tar_load(readme_pipeline_mermaid)
    cat(readme_pipeline_mermaid)
  }, error = function(e) {
    # Fallback to text description when target not built
    cat("1. **Data**: football-data.co.uk match results + Pinnacle closing odds\n")
    cat("2. **Features**: Rolling goals, Elo ratings, devigged implied probabilities\n")
    cat("3. **Models**: Poisson GLM baseline, Dixon-Coles, Elo-based\n")
    cat("4. **Evaluation**: Walk-forward CV with log-loss, Brier score, calibration\n")
    cat("5. **Decisions**: Fractional Kelly staking with drawdown guardrails\n")
  })
} else {
  cat("1. **Data**: football-data.co.uk match results + Pinnacle closing odds\n")
  cat("2. **Features**: Rolling goals, Elo ratings, devigged implied probabilities\n")
  cat("3. **Models**: Poisson GLM baseline, Dixon-Coles, Elo-based\n")
  cat("4. **Evaluation**: Walk-forward CV with log-loss, Brier score, calibration\n")
  cat("5. **Decisions**: Fractional Kelly staking with drawdown guardrails\n")
}
```

## Project Structure

```{r}
#| echo: false
#| eval: true
fs::dir_tree(recurse = 2)
```

## License

MIT
