---
title: "Data Cleaning & QC"
subtitle: "Automated quality control pipeline"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    page-layout: full
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Data Cleaning & QC}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(htmltools)
library(DT)
library(dplyr)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) {
    rds_path <- system.file(file.path("extdata", "vignettes", paste0(name, ".rds")), package = "footbet")
    if (!nzchar(rds_path) || !file.exists(rds_path))
      rds_path <- here::here("inst", "extdata", "vignettes", paste0(name, ".rds"))
    if (file.exists(rds_path)) result <- readRDS(rds_path)
  }
  if (is.null(result)) {
    htmltools::div(style = "background: #dc3545; color: white; padding: 1em; border-radius: 4px;",
      htmltools::p(paste0("[MISSING EVIDENCE] Target `", name, "` not found.")))
  } else result
}
```

Part 2 of the [footbet](https://johngavin.github.io/footbet/) analytics guide.
See also: [Data Sources](data-sources.html) |
[EDA](eda.html) | [Models & Betting](models-betting.html) |
[Glossary](glossary.html)

The footbet pipeline includes automated quality control checks that run after
data parsing. This section summarises the QC outputs.

## Match Completeness

The chart below shows the percentage of matches per league-season with a
valid full-time result (FTR). A value below 100% indicates matches where the
result is missing --- typically from ongoing or postponed fixtures in the current
season.

```{r vig-completeness-plot}
safe_tar_read("vig_completeness_plot")
```

::: {.callout-note}
**Key Finding:** Most historical league-seasons achieve 100% completeness.
Current season shows partial data (matches not yet played).
:::

## Pinnacle Coverage

Pinnacle odds are the primary benchmark for model evaluation. The chart
shows the percentage of matches with all three Pinnacle 1X2 closing odds
present.

Coverage gaps arise from:

- **Division 2 leagues**: Pinnacle may not offer odds for lower divisions
- **Feed breaks**: The football-data.co.uk Pinnacle feed experienced a break
  from mid-2025
- **Early seasons**: Pinnacle data availability improves in more recent seasons

```{r vig-pinnacle-coverage-plot}
safe_tar_read("vig_pinnacle_coverage_plot")
```

::: {.callout-note}
**Key Finding:** Top divisions have near-100% coverage from ~2016 onward.
Division 2 coverage is variable. 0% coverage indicates feed breaks or no
Pinnacle market.
:::

## Missing Data

The chart below shows the percentage of missing values for each
column across all parsed matches. Match statistics (shots, corners, cards) are
more frequently missing than core fields.

```{r vig-missing-data-by-column}
safe_tar_read("vig_missing_data_heatmap")
```

The table below provides detailed missing counts with variable descriptions.
See [Glossary > Data Terminology](glossary.html#data-terminology) for full definitions.

```{r vig-missing-data-table}
missing_data <- safe_tar_read("vig_missing_data_by_column")
if (is.data.frame(missing_data)) {
  missing_data |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Missing data by column. Match statistics often unavailable in older seasons."
      ),
      filter = "top",
      options = list(pageLength = 15, dom = "Bfrtip")
    )
} else {
  missing_data
}
```

## Anomaly Detection

The pipeline flags matches with:

- **Extreme scorelines**: Total goals > 10 (very rare, may indicate data entry errors)
- **Missing dates**: Matches without a parseable date
- **Future matches**: Dates more than 7 days beyond the pipeline run date
- **Missing team names**: Rows with NA home or away team

```{r vig-anomalies-table}
anomalies <- safe_tar_read("vig_anomalies_table")
if (is.data.frame(anomalies) && nrow(anomalies) > 0) {
  anomalies |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Flagged anomalies for manual review. Empty table = no anomalies detected."
      ),
      filter = "top",
      options = list(pageLength = 10, dom = "Bfrtip")
    )
} else if (is.data.frame(anomalies)) {
  htmltools::div(
    style = "background: #28a745; color: white; padding: 1em; border-radius: 4px;",
    "No anomalies detected in the dataset."
  )
} else {
  anomalies
}
```

## QC Summary

The `qc_summary` target combines all QC checks into a single report. The
pipeline computes total matches parsed, odds rows, completeness percentage,
Pinnacle coverage, and anomaly count. These targets serve as data
contracts --- downstream model targets depend on QC to ensure data quality.
