---
title: "Data Sources & Coverage"
subtitle: "Football-data.co.uk CSV pipeline"
format:
  html:
    theme: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    page-layout: full
    code-fold: true
    include-in-header:
      text: |
        <script>document.documentElement.setAttribute('data-bs-theme','dark');</script>
        <style>html,body{background:#000!important;color:#fff!important;}</style>
        <script src="https://cdn.jsdelivr.net/npm/mermaid/dist/mermaid.min.js"></script>
        <script>
          mermaid.initialize({
            startOnLoad: true,
            securityLevel: 'loose',
            theme: 'dark',
            themeVariables: {
              background: '#000000',
              primaryColor: '#999999',
              primaryTextColor: '#000000',
              primaryBorderColor: '#CC0000',
              lineColor: '#CC0000',
              secondaryColor: '#333333',
              tertiaryColor: '#333333'
            }
          });
        </script>
vignette: >
  %\VignetteIndexEntry{Data Sources & Coverage}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE
)
library(targets)
library(htmltools)
library(DT)
library(dplyr)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) {
    rds_path <- system.file(
      file.path("extdata", "vignettes", paste0(name, ".rds")),
      package = "footbet"
    )
    if (!nzchar(rds_path) || !file.exists(rds_path)) {
      rds_path <- here::here("inst", "extdata", "vignettes", paste0(name, ".rds"))
    }
    if (file.exists(rds_path)) result <- readRDS(rds_path)
  }
  if (is.null(result)) {
    htmltools::div(
      style = "background: #dc3545; color: white; padding: 1em; border-radius: 4px;",
      htmltools::p(paste0("[MISSING EVIDENCE] Target `", name, "` not found.")),
      htmltools::p("Run `targets::tar_make()` to build vignette outputs.",
                   style = "font-weight: normal; font-size: 0.9em;")
    )
  } else {
    result
  }
}

dark_dt <- function(data, caption = "", ...) {
  is_num <- sapply(data, function(x) is.numeric(x) && !is.list(x))
  num_cols <- names(data)[is_num]
  int_cols <- if (length(num_cols) > 0) {
    num_cols[sapply(data[num_cols], function(x) all(x == floor(x), na.rm = TRUE))]
  } else character(0)
  score_cols <- setdiff(num_cols, int_cols)
  dt <- DT::datatable(data, caption = caption,
    options = list(pageLength = 20, dom = "t", scrollX = TRUE, autoWidth = TRUE,
      initComplete = DT::JS(
        "function(settings, json) {",
        "  $(this.api().table().container()).css({'background-color': '#000000', 'color': '#ffffff'});",
        "  $(this.api().table().header()).css({'background-color': '#000000', 'color': '#ffffff', 'border-bottom': '1px solid #444'});",
        "  $('td', this.api().table().body()).css({'background-color': '#000000', 'color': '#ffffff', 'border-bottom': '1px solid #222'});",
        "}")),
    escape = FALSE, rownames = FALSE, ...)
  if (length(score_cols) > 0) dt <- DT::formatRound(dt, columns = score_cols, digits = 3)
  if (length(int_cols) > 0) dt <- DT::formatRound(dt, columns = int_cols, digits = 0)
  dt
}
```

```{=html}
<link rel="stylesheet" href="https://cdn.datatables.net/1.13.8/css/jquery.dataTables.min.css">
<script src="https://cdn.datatables.net/1.13.8/js/jquery.dataTables.min.js"></script>
<script>
// Force dark mode for this article
document.documentElement.setAttribute('data-bs-theme', 'dark');
</script>
<style>
/* Force dark page background */
body, .page-body, main, article, .contents { background-color: #000000 !important; color: #ffffff !important; }
/* Dark tables: black bg, white text, horizontal scroll */
table, .table, table.dataTable { background-color: #000000 !important; color: #ffffff !important; }
table td, table th, .table td, .table th { color: #ffffff !important; background-color: #000000 !important; border-color: #333 !important; }
table thead th, .table thead th { color: #ffffff !important; font-weight: bold !important; background-color: #111 !important; border-color: #333 !important; }
table tbody tr:hover td { background-color: #222 !important; }
.dataTables_info, .dataTables_paginate, .dataTables_length,
.dataTables_filter label, .dataTables_wrapper { color: #ffffff !important; }
.dataTables_filter input, .dataTables_length select {
  color: #ffffff !important; background-color: #333 !important; border-color: #555 !important; }
caption, .dataTables_wrapper caption { color: #cccccc !important; }
/* Horizontal scroll for wide tables */
.dataTables_wrapper, .cell-output-display { overflow-x: auto !important; }
table { white-space: nowrap; }
/* Ensure all text is white on dark */
p, li, h1, h2, h3, h4, h5, h6, span, em, strong, a, code { color: #ffffff !important; }
a:hover { color: #66b3ff !important; }
code { background-color: #222 !important; }
.callout { background-color: #111 !important; border-color: #333 !important; }
</style>
```

Part 1 of the [footbet](https://johngavin.github.io/footbet/) analytics guide.
See also: [Data Cleaning](data-cleaning.html) |
[EDA](eda.html) | [Models & Betting](models-betting.html) |
[Glossary](glossary.html)

The [footbet](https://johngavin.github.io/footbet/) package downloads match
data from [football-data.co.uk](https://www.football-data.co.uk/), a free
source of European football results and bookmaker odds maintained since the
mid-1990s. Each CSV file contains one league-season and includes full-time and
half-time scores, match statistics (shots, corners, cards), and closing odds
from multiple bookmakers including
[Pinnacle](https://en.wikipedia.org/wiki/Pinnacle_(sportsbook)) --- widely
regarded as the sharpest bookmaker in the industry.

The URL pattern is `https://www.football-data.co.uk/mmz4281/{season}/{league}.csv`,
where `{season}` is a 4-digit code (e.g., `2324` for 2023-24) and `{league}`
is a 2-character code (e.g., `E0` for the English Premier League).

`footbet::target_leagues()` returns 10 leagues covering the top 2 divisions
of the "big five" European football countries.

## Data Pipeline Overview

```{r vig-pipeline-diagram}
#| echo: false
#| output: asis
safe_tar_read("vig_pipeline_html")
```

*Data pipeline flowchart showing the end-to-end processing stages. Nodes represent pipeline steps: CSV download, parsing via `parse_fd_csv()`, odds extraction via `parse_fd_odds()`, quality control, and model fitting. Red arrows indicate data flow direction. The pipeline is implemented as a targets DAG (directed acyclic graph) ensuring reproducibility and incremental builds. Source: [footbet](https://johngavin.github.io/footbet/) package architecture. See [Glossary > Data Terminology](glossary.html#data-terminology).*

## Data Coverage

```{r vig-data-source-summary}
data_summary <- safe_tar_read("vig_data_source_summary")
if (is.data.frame(data_summary)) {
  dark_dt(data_summary,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Data coverage by league. 10 leagues across 5 countries, top 2 divisions each."
    )
  )
} else {
  data_summary
}
```

*Data coverage summary table. Columns: league code (e.g., E0 = English Premier League), country, division, number of seasons available, total matches parsed. The dataset covers 10 leagues across England, Germany, Italy, Spain, and France, with top two divisions each. Division 1 leagues typically span more seasons than Division 2 due to longer data availability on football-data.co.uk. Total match counts range from several thousand for well-covered leagues to fewer for recently added ones. Source: [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Data Terminology](glossary.html#data-terminology).*

## Coverage Grid

The grid below shows the number of parsed matches per league-season cell.
Zeroes indicate seasons where the CSV was unavailable or empty.

```{r vig-league-season-grid}
grid <- safe_tar_read("vig_league_season_grid")
if (is.data.frame(grid)) {
  dark_dt(grid,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Matches per league-season. 0 = data unavailable. ~380 for Div 1, ~460+ for Div 2."
    )
  )
} else {
  grid
}
```

*League-season coverage grid. Rows: league codes (E0, E1, D1, D2, I1, I2, SP1, SP2, F1, F2). Columns: season codes (e.g., 2324). Cell values: number of matches parsed for that league-season combination. Zero values indicate missing or empty CSV files. Division 1 leagues (E0, D1, I1, SP1, F1) have approximately 380 matches per season (20 teams, 38 rounds). Division 2 leagues have more matches (460-550) due to larger league sizes (24 teams). Coverage is most complete for recent seasons. Source: [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Data Terminology](glossary.html#data-terminology).*

```{r vig-matches-per-season-plot}
safe_tar_read("vig_matches_per_season_plot")
```

*Matches per season by league over time. X-axis: season (year). Y-axis: number of matches. Lines/bars: one per league, colour-coded by league code. Division 1 leagues show a stable ~380 matches per season. Division 2 leagues show higher counts (~460-550) reflecting larger league sizes. Gaps indicate seasons where data was unavailable. The plot confirms consistent data collection for recent seasons across all 10 leagues. Source: [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Data Terminology](glossary.html#data-terminology).*

::: {.callout-note}
**Key Finding:** Division 1 leagues (E0, D1, I1, SP1, F1) have ~380
matches/season. Division 2 leagues have more (~460-550 matches/season).
:::

## Pinnacle Odds

Not all leagues and seasons have Pinnacle odds. The table below shows the
percentage of matches with complete Pinnacle 1X2 closing odds (PSH, PSD, PSA)
and over/under odds.

```{r vig-odds-columns-available}
odds_avail <- safe_tar_read("vig_odds_columns_available")
if (is.data.frame(odds_avail)) {
  dark_dt(odds_avail,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Pinnacle odds coverage. pct_1x2 = % with home/draw/away odds. pct_over_under = % with O/U 2.5."
    )
  )
} else {
  odds_avail
}
```

*Pinnacle odds availability by league and season. Columns: league code, season, pct_1x2 (percentage of matches with complete home/draw/away Pinnacle closing odds), pct_over_under (percentage with over/under 2.5 goals odds). Top-division leagues achieve near-100% Pinnacle coverage from approximately 2016 onward. Division 2 leagues show variable coverage --- Pinnacle may not offer markets for lower divisions. Coverage gaps also arise from feed interruptions on football-data.co.uk. This coverage determines which matches can be used for model evaluation against the sharp-market benchmark. Source: [football-data.co.uk](https://www.football-data.co.uk/) Pinnacle feed. See [Glossary > Implied Probability](glossary.html) and [Glossary > Overround](glossary.html).*

## Column Schema

After parsing, `footbet::parse_fd_csv()` produces a standardised tibble with
the following columns:

| Column | Description | Type |
|--------|-------------|------|
| `match_id` | Deterministic ID: `league_date_home_away` | character |
| `league_code` | 2-char league code (e.g., E0, D1) | character |
| `season` | 4-digit season code (e.g., 2324) | character |
| `match_date` | Match date | Date |
| `home_team` / `away_team` | Team names | character |
| `fthg` / `ftag` | Full-time home/away goals | integer |
| `ftr` | Full-time result: H, D, or A | character |
| `hthg` / `htag` | Half-time home/away goals | integer |
| `htr` | Half-time result | character |
| `hs` / `as_` | Home/away shots | integer |
| `hst` / `ast` | Home/away shots on target | integer |
| `hf` / `af` | Home/away fouls | integer |
| `hc` / `ac` | Home/away corners | integer |
| `hy` / `ay` | Home/away yellow cards | integer |
| `hr` / `ar` | Home/away red cards | integer |

: Columns produced by `parse_fd_csv()`. See [Glossary](glossary.html#data-terminology) for definitions. {.striped}

`footbet::parse_fd_odds()` extracts Pinnacle closing odds:

| Column | Description |
|--------|-------------|
| `psh` / `psd` / `psa` | Pinnacle closing 1X2 odds (Home/Draw/Away) |
| `p_over25` / `p_under25` | Pinnacle over/under 2.5 goals odds |

: Columns produced by `parse_fd_odds()`. Pinnacle odds serve as the sharp-market benchmark. {.striped}

```{r build-info, echo = FALSE, results = "asis"}
safe_tar_read("vig_build_info")
```
