---
title: "Data Sources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 6
)
library(targets)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) invisible(NULL) else result
}
```

## Football-data.co.uk

The [footbet](https://johngavin.github.io/footbet/) package downloads match
data from [football-data.co.uk](https://www.football-data.co.uk/), a free
source of European football results and bookmaker odds maintained since the
mid-1990s. Each CSV file contains one league-season and includes full-time and
half-time scores, match statistics (shots, corners, cards), and closing odds
from multiple bookmakers including
[Pinnacle](https://en.wikipedia.org/wiki/Pinnacle_(sportsbook)) — widely
regarded as the sharpest bookmaker in the industry.

The URL pattern is `https://www.football-data.co.uk/mmz4281/{season}/{league}.csv`,
where `{season}` is a 4-digit code (e.g., `2324` for 2023-24) and `{league}`
is a 2-character code (e.g., `E0` for the English Premier League).

`footbet::target_leagues()` returns 10 leagues covering the top 2 divisions
of the "big five" European football countries.

```{r vig-data-source-summary}
safe_tar_read("vig_data_source_summary")
```

## League & Season Coverage

The grid below shows the number of parsed matches per league-season cell.
Zeroes indicate seasons where the CSV was unavailable or empty.

```{r vig-league-season-grid}
safe_tar_read("vig_league_season_grid")
```

```{r vig-matches-per-season-plot, fig.cap = "Match count by league and season. Bars show total parsed matches. Key: Division 1 leagues (E0, D1, I1, SP1, F1) have ~380 matches/season; Division 2 leagues have more (~460-550). Source: football-data.co.uk."}
safe_tar_read("vig_matches_per_season_plot")
```

## Pinnacle Odds Availability

Not all leagues and seasons have Pinnacle odds. The table below shows the
percentage of matches with complete Pinnacle 1X2 closing odds (PSH, PSD, PSA)
and over/under odds.

Pinnacle coverage dropped significantly for some leagues from mid-2025 due to
a feed break at football-data.co.uk.

```{r vig-odds-columns-available}
safe_tar_read("vig_odds_columns_available")
```

## Column Schema

After parsing, `footbet::parse_fd_csv()` produces a standardised tibble with
the following columns:

| Column | Description | Type |
|--------|-------------|------|
| `match_id` | Deterministic ID: `league_date_home_away` | character |
| `league_code` | 2-char league code (e.g., E0, D1) | character |
| `season` | 4-digit season code (e.g., 2324) | character |
| `match_date` | Match date | Date |
| `home_team` / `away_team` | Team names | character |
| `fthg` / `ftag` | Full-time home/away goals | integer |
| `ftr` | Full-time result: H, D, or A | character |
| `hthg` / `htag` | Half-time home/away goals | integer |
| `htr` | Half-time result | character |
| `hs` / `as_` | Home/away shots | integer |
| `hst` / `ast` | Home/away shots on target | integer |
| `hf` / `af` | Home/away fouls | integer |
| `hc` / `ac` | Home/away corners | integer |
| `hy` / `ay` | Home/away yellow cards | integer |
| `hr` / `ar` | Home/away red cards | integer |

`footbet::parse_fd_odds()` extracts Pinnacle closing odds:

| Column | Description |
|--------|-------------|
| `psh` / `psd` / `psa` | Pinnacle closing 1X2 odds (Home/Draw/Away) |
| `p_over25` / `p_under25` | Pinnacle over/under 2.5 goals odds |
