---
title: "Data Cleaning & Quality Control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Cleaning & Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 6
)
library(targets)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) invisible(NULL) else result
}
```

The footbet pipeline includes automated quality control checks that run after
data parsing. This vignette summarises the QC outputs.

## Match Completeness

The heatmap below shows the percentage of matches per league-season with a
valid full-time result (FTR). A value below 100% indicates matches where the
result is missing — typically from ongoing or postponed fixtures in the current
season.

```{r vig-completeness-plot, fig.cap = "Match result completeness heatmap. Fill = percentage of matches with non-missing FTR. Key: Most historical league-seasons achieve 100%; current season shows partial data. Source: football-data.co.uk, parsed by footbet::parse_fd_csv()."}
safe_tar_read("vig_completeness_plot")
```

## Pinnacle Odds Coverage

Pinnacle odds are the primary benchmark for model evaluation. The heatmap
shows the percentage of matches with all three Pinnacle 1X2 closing odds
present.

Coverage gaps arise from:

- **Division 2 leagues**: Pinnacle may not offer odds for lower divisions in
  some countries.
- **Feed breaks**: The football-data.co.uk Pinnacle feed experienced a break
  from mid-2025, causing 0% coverage for affected seasons.
- **Early seasons**: Pinnacle data availability improves in more recent
  seasons.

```{r vig-pinnacle-coverage-plot, fig.cap = "Pinnacle 1X2 odds coverage heatmap. Fill = percentage of matches with complete PSH/PSD/PSA. Key: Top divisions have near-100% coverage from ~2016 onward; Division 2 coverage is variable. Source: football-data.co.uk Pinnacle columns."}
safe_tar_read("vig_pinnacle_coverage_plot")
```

## Anomaly Detection

The pipeline flags matches with:

- **Extreme scorelines**: Total goals > 10 (very rare, may indicate data
  entry errors)
- **Missing dates**: Matches without a parseable date
- **Future matches**: Dates more than 7 days beyond the pipeline run date
- **Missing team names**: Rows with NA home or away team

```{r vig-anomalies-table}
safe_tar_read("vig_anomalies_table")
```

## Missing Data Profile

The table below shows the count and percentage of missing values for each
column across all parsed matches. Match statistics (shots, corners, cards) are
more frequently missing than core fields (teams, dates, scores), especially
for older seasons or lower divisions.

```{r vig-missing-data-by-column}
safe_tar_read("vig_missing_data_by_column")
```

## Data Quality Summary

The `qc_summary` target combines all QC checks into a single report with a
timestamp. The pipeline computes:

- **Total matches parsed**: Row count across all leagues/seasons
- **Total odds rows**: Pinnacle odds records parsed
- **Completeness**: Percentage of matches with valid results
- **Pinnacle coverage**: Percentage with 1X2 odds
- **Anomaly count**: Flagged rows for manual review

The QC targets serve as data contracts — downstream model targets depend on
`qc_summary` to ensure data quality issues are surfaced before model fitting.
