---
title: "Data Cleaning & QC"
subtitle: "Automated quality control pipeline"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    page-layout: full
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Data Cleaning & QC}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(htmltools)
library(DT)
library(dplyr)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) {
    rds_path <- system.file(file.path("extdata", "vignettes", paste0(name, ".rds")), package = "footbet")
    if (!nzchar(rds_path) || !file.exists(rds_path))
      rds_path <- here::here("inst", "extdata", "vignettes", paste0(name, ".rds"))
    if (file.exists(rds_path)) result <- readRDS(rds_path)
  }
  if (is.null(result)) {
    htmltools::div(style = "background: #dc3545; color: white; padding: 1em; border-radius: 4px;",
      htmltools::p(paste0("[MISSING EVIDENCE] Target `", name, "` not found.")))
  } else result
}
```

```{=html}
<style>
table.dataTable tbody td, table.dataTable tbody th,
.table tbody td, .table tbody th { color: #cccccc !important; }
table.dataTable thead th, table.dataTable thead td,
.table thead th, .table thead td { color: #ffffff !important; }
.dataTables_info, .dataTables_paginate, .dataTables_length,
.dataTables_filter label { color: #cccccc !important; }
.dataTables_filter input, .dataTables_length select {
  color: #cccccc !important; background-color: #333 !important; }
</style>
```

Part 2 of the [footbet](https://johngavin.github.io/footbet/) analytics guide.
See also: [Data Sources](data-sources.html) |
[EDA](eda.html) | [Models & Betting](models-betting.html) |
[Glossary](glossary.html)

The footbet pipeline includes automated quality control checks that run after
data parsing. This section summarises the QC outputs.

## Match Completeness

The chart below shows the percentage of matches per league-season with a
valid full-time result (FTR). A value below 100% indicates matches where the
result is missing --- typically from ongoing or postponed fixtures in the current
season.

```{r vig-completeness-plot}
safe_tar_read("vig_completeness_plot")
```

*Match completeness by league-season. X-axis: season (year). Y-axis: percentage of matches with a valid full-time result (FTR). Colour = league code (E0, D1, etc.). A value of 100% means all matches in that league-season have a recorded result. Most historical seasons achieve full completeness. The current season shows partial data because not all fixtures have been played yet. This metric is critical for ensuring model training data does not include unresolved matches. Source: [football-data.co.uk](https://www.football-data.co.uk/) parsed via `parse_fd_csv()`. See [Glossary > FTR](glossary.html#data-terminology).*

::: {.callout-note}
**Key Finding:** Most historical league-seasons achieve 100% completeness.
Current season shows partial data (matches not yet played).
:::

## Pinnacle Coverage

Pinnacle odds are the primary benchmark for model evaluation. The chart
shows the percentage of matches with all three Pinnacle 1X2 closing odds
present.

Coverage gaps arise from:

- **Division 2 leagues**: Pinnacle may not offer odds for lower divisions
- **Feed breaks**: The football-data.co.uk Pinnacle feed experienced a break
  from mid-2025
- **Early seasons**: Pinnacle data availability improves in more recent seasons

```{r vig-pinnacle-coverage-plot}
safe_tar_read("vig_pinnacle_coverage_plot")
```

*Pinnacle 1X2 odds coverage by league-season. X-axis: season (year). Y-axis: percentage of matches with complete Pinnacle home/draw/away closing odds (PSH, PSD, PSA). Colour = league code. Top-division leagues (E0, D1, I1, SP1, F1) achieve near-100% coverage from approximately 2016 onward. Division 2 leagues (E1, D2, I2, SP2, F2) show variable and often lower coverage, as Pinnacle may not offer markets for these competitions. Coverage of 0% indicates complete feed breaks or no Pinnacle market for that league-season. Source: [football-data.co.uk](https://www.football-data.co.uk/) Pinnacle feed. See [Glossary > Implied Probability](glossary.html) and [Glossary > Sharp Bookmaker](glossary.html).*

::: {.callout-note}
**Key Finding:** Top divisions have near-100% coverage from ~2016 onward.
Division 2 coverage is variable. 0% coverage indicates feed breaks or no
Pinnacle market.
:::

## Missing Data

The chart below shows the percentage of missing values for each
column across all parsed matches. Match statistics (shots, corners, cards) are
more frequently missing than core fields.

```{r vig-missing-data-by-column}
safe_tar_read("vig_missing_data_heatmap")
```

*Missing data heatmap by column and league. X-axis: column name (fthg, ftag, ftr, hs, as_, hst, ast, hc, ac, hf, af, hy, ay, hr, ar). Y-axis: league code. Cell colour intensity = percentage of missing values (darker = more missing). Core result fields (fthg, ftag, ftr) have near-zero missingness. Match statistics columns (shots, corners, cards) show higher missingness, especially in older seasons and Division 2 leagues where football-data.co.uk may not collect detailed statistics. This pattern informs which features are reliably available for modelling. Source: [football-data.co.uk](https://www.football-data.co.uk/) parsed data. See [Glossary > Match Statistics](glossary.html#match-statistics).*

The table below provides detailed missing counts with variable descriptions.
See [Glossary > Data Terminology](glossary.html#data-terminology) for full definitions.

```{r vig-missing-data-table}
missing_data <- safe_tar_read("vig_missing_data_by_column")
if (is.data.frame(missing_data)) {
  missing_data |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Missing data by column. Match statistics often unavailable in older seasons."
      ),
      filter = "top",
      options = list(pageLength = 15, dom = "Bfrtip")
    )
} else {
  missing_data
}
```

*Missing data counts by variable. Columns: variable name, description, number of missing values, percentage missing, leagues most affected. Core fields (match_date, home_team, away_team, fthg, ftag, ftr) are nearly always present. Match statistics (hs, hst, hc, hf, hy, hr and their away equivalents) have higher missingness, particularly in seasons before detailed statistics were collected. Pinnacle odds columns (psh, psd, psa) are missing for league-seasons without Pinnacle coverage. Source: [football-data.co.uk](https://www.football-data.co.uk/) parsed data. See [Glossary > Match Statistics](glossary.html#match-statistics) and [Glossary > Data Terminology](glossary.html#data-terminology).*

## Anomaly Detection

The pipeline flags matches with:

- **Extreme scorelines**: Total goals > 10 (very rare, may indicate data entry errors)
- **Missing dates**: Matches without a parseable date
- **Future matches**: Dates more than 7 days beyond the pipeline run date
- **Missing team names**: Rows with NA home or away team

```{r vig-anomalies-table}
anomalies <- safe_tar_read("vig_anomalies_table")
if (is.data.frame(anomalies) && nrow(anomalies) > 0) {
  anomalies |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Flagged anomalies for manual review. Empty table = no anomalies detected."
      ),
      filter = "top",
      options = list(pageLength = 10, dom = "Bfrtip")
    )
} else if (is.data.frame(anomalies)) {
  htmltools::div(
    style = "background: #28a745; color: white; padding: 1em; border-radius: 4px;",
    "No anomalies detected in the dataset."
  )
} else {
  anomalies
}
```

*Anomaly detection results table. Columns: match_id, league_code, season, match_date, home_team, away_team, fthg, ftag, anomaly_type (extreme_score / missing_date / future_date / missing_team). Each row is a flagged match requiring manual review. Extreme scorelines (>10 total goals) are rare but may indicate data entry errors. An empty table (green banner) means no anomalies were detected, confirming data integrity. Source: automated QC pipeline applied to [football-data.co.uk](https://www.football-data.co.uk/) data. See [Glossary > FTHG / FTAG](glossary.html#data-terminology).*

## QC Summary

The `qc_summary` target combines all QC checks into a single report. The
pipeline computes total matches parsed, odds rows, completeness percentage,
Pinnacle coverage, and anomaly count. These targets serve as data
contracts --- downstream model targets depend on QC to ensure data quality.
