---
title: "Data Cleaning & Quality Control"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: false
    page-layout: full
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Data Cleaning & Quality Control}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE
)
library(targets)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) invisible(NULL) else result
}
```

The footbet pipeline includes automated quality control checks that run after
data parsing. This vignette summarises the QC outputs.

::: {.panel-tabset}

## Match Completeness

The heatmap below shows the percentage of matches per league-season with a
valid full-time result (FTR). A value below 100% indicates matches where the
result is missing — typically from ongoing or postponed fixtures in the current
season.

```{r vig-completeness-plot}
safe_tar_read("vig_completeness_plot")
```

**Key findings:**

- Most historical league-seasons achieve 100% completeness
- Current season shows partial data (matches not yet played)
- Greyed cells indicate <80% completeness or missing data

## Pinnacle Coverage

Pinnacle odds are the primary benchmark for model evaluation. The heatmap
shows the percentage of matches with all three Pinnacle 1X2 closing odds
present.

Coverage gaps arise from:

- **Division 2 leagues**: Pinnacle may not offer odds for lower divisions
- **Feed breaks**: The football-data.co.uk Pinnacle feed experienced a break
  from mid-2025
- **Early seasons**: Pinnacle data availability improves in more recent seasons

```{r vig-pinnacle-coverage-plot}
safe_tar_read("vig_pinnacle_coverage_plot")
```

**Key findings:**

- Top divisions have near-100% coverage from ~2016 onward
- Division 2 coverage is variable
- 0% coverage indicates feed breaks or no Pinnacle market

## Missing Data

The chart below shows the percentage of missing values for each
column across all parsed matches. Match statistics (shots, corners, cards) are
more frequently missing than core fields.

```{r vig-missing-data-by-column}
safe_tar_read("vig_missing_data_heatmap")
```

The table below provides detailed missing counts:

```{r vig-missing-data-table}
safe_tar_read("vig_missing_data_by_column")
```

## Anomaly Detection

The pipeline flags matches with:

- **Extreme scorelines**: Total goals > 10 (very rare, may indicate data
  entry errors)
- **Missing dates**: Matches without a parseable date
- **Future matches**: Dates more than 7 days beyond the pipeline run date
- **Missing team names**: Rows with NA home or away team

```{r vig-anomalies-table}
safe_tar_read("vig_anomalies_table")
```

## QC Summary

The `qc_summary` target combines all QC checks into a single report with a
timestamp. The pipeline computes:

- **Total matches parsed**: Row count across all leagues/seasons
- **Total odds rows**: Pinnacle odds records parsed
- **Completeness**: Percentage of matches with valid results
- **Pinnacle coverage**: Percentage with 1X2 odds
- **Anomaly count**: Flagged rows for manual review

The QC targets serve as data contracts — downstream model targets depend on
`qc_summary` to ensure data quality issues are surfaced before model fitting.

:::
