---
title: "Exploratory Data Analysis"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: false
    page-layout: full
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Exploratory Data Analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE
)
library(targets)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) invisible(NULL) else result
}
```

This vignette explores key statistical patterns in European football match
data across 10 leagues and multiple seasons. All visualizations are interactive
with dark theme styling.

::: {.panel-tabset}

## Goal Distributions

Football goals are often modelled as
[Poisson-distributed](https://en.wikipedia.org/wiki/Poisson_distribution)
random variables. The chart below compares the observed goal distribution
against a Poisson overlay for home and away teams.

```{r vig-goals-distribution}
safe_tar_read("vig_goals_distribution")
```

**Key findings:**

- Home teams score slightly more than away teams (home advantage)
- The Poisson distribution is a reasonable first approximation
- The model underpredicts 0-0 draws and some high-scoring outcomes
- See Dixon & Coles (1997) for the correlation correction term

## Result Proportions

Home advantage is a well-documented phenomenon in football. The chart below
shows the proportion of Home wins, Draws, and Away wins by league, compared
to a uniform 33/33/33 baseline.

```{r vig-result-proportions}
safe_tar_read("vig_result_proportions")
```

**Key findings:**

- Home win rate ranges 43-47% across leagues
- Draws: 25-28%
- Away wins: 27-30%
- All leagues show significant home advantage

## Home Advantage Trends

Home advantage has declined in recent years, with a notable drop during the
COVID-19 pandemic (2020-21 season) when matches were played in empty stadiums.

```{r vig-home-advantage-by-league}
safe_tar_read("vig_home_advantage_by_league")
```

**Key findings:**

- COVID seasons (2020-21) show reduced home advantage across all leagues
- Post-COVID recovery varies by league
- Long-term declining trend visible in most leagues

## Common Scorelines

The heatmap below shows the frequency of each scoreline (home goals x away
goals). Hover over cells to see exact percentages.

```{r vig-scoreline-heatmap}
safe_tar_read("vig_scoreline_heatmap")
```

**Key findings:**

- 1-0 (~12%), 1-1 (~11%), 2-1 (~10%) are most common
- Scorelines above 4-4 are extremely rare (<0.5% combined)
- The distribution is right-skewed with a heavy tail

## Goals Per Season

Is scoring increasing over time? The chart below tracks mean total goals
per match by season and league. Use the range slider to zoom into specific periods.

```{r vig-goals-per-season-trend}
safe_tar_read("vig_goals_per_season_trend")
```

**Key findings:**

- Slight upward trend since 2015 in most leagues
- Bundesliga consistently among highest-scoring (~3.0 goals/match)
- Serie A shows notable increase in recent seasons

## Outliers

Matches with more than 7 total goals are rare outliers. These are worth
inspecting for data quality and as edge cases for model robustness.

```{r vig-outlier-matches}
safe_tar_read("vig_outlier_matches")
```

## Match Statistics

Median match statistics (shots, corners, fouls, cards) by league. These vary
across leagues reflecting different playing styles.

```{r vig-typical-match-stats}
safe_tar_read("vig_typical_match_stats")
```

## Pinnacle Overround

The [overround](https://en.wikipedia.org/wiki/Vigorish) (or vig) is the
bookmaker's built-in margin. Pinnacle's margin (~2-4%) is the lowest in the
industry, making their odds the best available benchmark for "true"
probabilities.

```{r vig-overround-by-league}
safe_tar_read("vig_overround_by_league")
```

**Key findings:**

- Pinnacle margin is ~2-4%, far below soft bookmakers (5-15%)
- Lower margin = fairer odds = better model benchmark
- Consistent across all leagues

## Pinnacle Calibration

A well-calibrated bookmaker's implied probabilities should match actual outcome
frequencies. The plot below bins devigged Pinnacle probabilities and compares
them to observed outcome rates.

```{r vig-odds-vs-result}
safe_tar_read("vig_odds_vs_result")
```

**Key findings:**

- Pinnacle is near-perfectly calibrated (points near diagonal)
- Slight favourite-longshot bias at extremes
- Size indicates sample size per bin

## Elo Ratings

[Elo ratings](https://en.wikipedia.org/wiki/Elo_rating_system) quantify
relative team strength. The chart below shows the distribution of final Elo
ratings per league.

```{r vig-elo-spread-plot}
safe_tar_read("vig_elo_spread_plot")
```

**Key findings:**

- Top divisions have wider Elo spreads (more competitive imbalance)
- Division 2 leagues show more compressed rating distributions
- Initial rating = 1500 for all teams

:::
