---
title: "Exploratory Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploratory Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 6
)
library(targets)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) invisible(NULL) else result
}
```

This vignette explores the key statistical patterns in European football match
data across 10 leagues and multiple seasons.

## Goal Distributions

Football goals are often modelled as
[Poisson-distributed](https://en.wikipedia.org/wiki/Poisson_distribution)
random variables. The histogram below compares the observed goal distribution
against a Poisson overlay, faceted by home and away teams.

```{r vig-goals-distribution, fig.cap = "Goal distribution: observed (bars) vs Poisson (orange line). Home teams score slightly more than away. The Poisson is a reasonable first approximation but underpredicts 0-0 draws. Source: football-data.co.uk, all leagues/seasons."}
safe_tar_read("vig_goals_distribution")
```

## Result Proportions

Home advantage is a well-documented phenomenon in football. The stacked bar
chart below shows the proportion of Home wins, Draws, and Away wins by league,
compared to a uniform 33/33/33 baseline.

```{r vig-result-proportions, fig.cap = "Result proportions by league. Stacked bars = H/D/A %. Dashed lines = uniform 33.3% baseline. Key: Home win rate ranges 43-47% across leagues; draws 25-28%; away wins 27-30%. Source: football-data.co.uk."}
safe_tar_read("vig_result_proportions")
```

## Home Advantage Trends

Home advantage has declined in recent years, with a notable drop during the
COVID-19 pandemic (2020-21 season) when matches were played in empty stadiums.

```{r vig-home-advantage-by-league, fig.cap = "Home win percentage by league over seasons. Orange dashed = overall mean. Key: COVID seasons (2020-21) show reduced home advantage; post-COVID recovery varies by league. Source: football-data.co.uk."}
safe_tar_read("vig_home_advantage_by_league")
```

## Common Scorelines

The heatmap below shows the frequency of each scoreline (home goals x away
goals). The most common results are 1-0, 1-1, and 2-1.

```{r vig-scoreline-heatmap, fig.cap = "Scoreline frequency heatmap (0-5 goals). Fill = % of all matches. Key: 1-0 (~12%), 1-1 (~11%), 2-1 (~10%) are most common. Scorelines >5 omitted. Source: football-data.co.uk, all leagues."}
safe_tar_read("vig_scoreline_heatmap")
```

## Goals per Season Trends

Is scoring increasing over time? The line chart below tracks mean total goals
per match by season, faceted by league.

```{r vig-goals-per-season-trend, fig.cap = "Mean total goals per match by season and league. Orange dashed = overall mean. Key: Slight upward trend since 2015 in most leagues; Bundesliga consistently among highest-scoring (~3.0 goals/match). Source: football-data.co.uk."}
safe_tar_read("vig_goals_per_season_trend")
```

## Outliers & Extremes

Matches with more than 7 total goals are rare outliers. These are worth
inspecting for data quality and as edge cases for model robustness.

```{r vig-outlier-matches}
safe_tar_read("vig_outlier_matches")
```

## Match Statistics Profile

Median match statistics (shots, corners, fouls, cards) by league. These vary
across leagues reflecting different playing styles.

```{r vig-typical-match-stats}
safe_tar_read("vig_typical_match_stats")
```

## Odds Market Structure

### Pinnacle Overround

The
[overround](https://en.wikipedia.org/wiki/Vigorish) (or vig) is the
bookmaker's built-in margin. Pinnacle's margin (~2-4%) is the lowest in the
industry, making their odds the best available benchmark for "true"
probabilities.

```{r vig-overround-by-league, fig.cap = "Pinnacle 1X2 overround by league. Box = IQR. Key: Pinnacle margin is ~2-4%, far below soft bookmakers (5-15%). Lower margin = fairer odds = better model benchmark. Source: football-data.co.uk Pinnacle closing odds."}
safe_tar_read("vig_overround_by_league")
```

### Pinnacle Calibration

A well-calibrated bookmaker's implied probabilities should match actual outcome
frequencies. The plot below bins devigged Pinnacle probabilities and compares
them to observed outcome rates.

```{r vig-odds-vs-result, fig.cap = "Pinnacle calibration plot. X = mean devigged implied probability per 5% bin; Y = actual outcome frequency. Diagonal = perfect calibration. Key: Pinnacle is near-perfectly calibrated; slight favourite-longshot bias at extremes. Source: Pinnacle odds devigged via Shin method."}
safe_tar_read("vig_odds_vs_result")
```

## Elo Rating Spreads

[Elo ratings](https://en.wikipedia.org/wiki/Elo_rating_system) quantify
relative team strength. The boxplot below shows the distribution of final Elo
ratings per league.

```{r vig-elo-spread-plot, fig.cap = "Elo rating distribution by league. Box = IQR, dashed = starting Elo (1500). Key: Top divisions have wider Elo spreads (more competitive imbalance) than Division 2. Source: footbet::compute_elo()."}
safe_tar_read("vig_elo_spread_plot")
```
