---
title: "Exploratory Data Analysis"
subtitle: "Statistical patterns in European football"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    page-layout: full
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Exploratory Data Analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(htmltools)
library(DT)
library(dplyr)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) {
    rds_path <- system.file(file.path("extdata", "vignettes", paste0(name, ".rds")), package = "footbet")
    if (!nzchar(rds_path) || !file.exists(rds_path))
      rds_path <- here::here("inst", "extdata", "vignettes", paste0(name, ".rds"))
    if (file.exists(rds_path)) result <- readRDS(rds_path)
  }
  if (is.null(result)) {
    htmltools::div(style = "background: #dc3545; color: white; padding: 1em; border-radius: 4px;",
      htmltools::p(paste0("[MISSING EVIDENCE] Target `", name, "` not found.")))
  } else result
}
```

Part 3 of the [footbet](https://johngavin.github.io/footbet/) analytics guide.
See also: [Data Sources](data-sources.html) |
[Data Cleaning](data-cleaning.html) | [Models & Betting](models-betting.html) |
[Glossary](glossary.html)

This section explores key statistical patterns in European football match
data across 10 leagues and multiple seasons.

## Goal Distributions

Football goals are often modelled as
[Poisson-distributed](https://en.wikipedia.org/wiki/Poisson_distribution)
random variables. The chart below compares the observed goal distribution
against a Poisson overlay for home and away teams.

```{r vig-goals-distribution}
safe_tar_read("vig_goals_distribution")
```

*Observed vs Poisson-fitted goal distributions for home and away teams. X-axis: number of goals per match (0, 1, 2, 3, 4, 5+). Y-axis: proportion of matches. Blue bars = observed frequency, orange line = Poisson distribution fit. Two panels: home goals (left) and away goals (right). Home teams have a higher mean (~1.5 goals) than away teams (~1.2 goals), reflecting home advantage. The Poisson fit is reasonable but slightly underpredicts 0-0 draws and overpredicts mid-range scores. Source: all parsed matches from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Poisson Regression](glossary.html).*

```{r vig-poisson-test}
safe_tar_read("vig_poisson_test")
```

*Goodness-of-fit test for the Poisson assumption. Columns: test statistic, p-value, conclusion. A chi-squared test compares observed goal frequencies against Poisson-expected frequencies. A significant p-value indicates the Poisson distribution is an imperfect fit, motivating the Dixon-Coles correction for low-scoring outcomes. Despite the statistical rejection, the Poisson remains a useful baseline model for practical prediction. Source: aggregated goal counts from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Poisson Regression](glossary.html) and [Glossary > Dixon-Coles Model](glossary.html).*

::: {.callout-note}
**Key Finding:** Home teams score slightly more than away teams (home
advantage). The Poisson distribution is a reasonable first approximation,
though it underpredicts 0-0 draws and some high-scoring outcomes.
:::

## Result Proportions

Home advantage is a well-documented phenomenon in football. The chart below
shows the proportion of Home wins, Draws, and Away wins by league, compared
to a uniform 33/33/33 baseline.

```{r vig-result-proportions}
safe_tar_read("vig_result_proportions")
```

*Result proportions by league. X-axis: league code (E0, E1, D1, D2, I1, I2, SP1, SP2, F1, F2). Y-axis: proportion (0-1). Stacked bars: green = Home wins, grey = Draws, red = Away wins. Dashed horizontal line at 0.333 = uniform baseline. Home win rates range 43-47% across all leagues, well above the 33% baseline. Draw rates are 25-28%, and away win rates are 27-30%. Serie A (I1) and La Liga (SP1) show among the highest home advantages. Source: all parsed matches from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > 1X2 (Match Result)](glossary.html) and [Glossary > FTR](glossary.html#data-terminology).*

::: {.callout-note}
**Key Finding:** Home win rate ranges 43-47% across leagues. Draws: 25-28%.
Away wins: 27-30%. All leagues show significant home advantage.
:::

## Home Advantage Trends

Home advantage has declined in recent years, with a notable drop during the
COVID-19 pandemic (2020-21 season) when matches were played in empty stadiums.

```{r vig-home-advantage-by-league}
safe_tar_read("vig_home_advantage_by_league")
```

*Home win percentage by league over time. X-axis: season (year). Y-axis: home win percentage. Lines: one per league, colour-coded by league code. Dashed horizontal line at 33% = no home advantage baseline. Home advantage has gradually declined from ~48% in 2010 to ~44% in recent seasons. The COVID-19 season (2020-21) shows a sharp drop across all leagues when matches were played behind closed doors. Post-COVID recovery varies --- some leagues have rebounded while others remain below pre-pandemic levels. Source: [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > 1X2 (Match Result)](glossary.html).*

```{r vig-home-trend-test}
safe_tar_read("vig_home_trend_test")
```

*Statistical test for declining home advantage trend. Columns: coefficient, standard error, t-statistic, p-value. A linear regression of home win percentage on season tests whether the long-term decline is statistically significant. A negative coefficient confirms the downward trend. The effect size is small but consistent across leagues. Source: season-level aggregation from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > 1X2 (Match Result)](glossary.html).*

::: {.callout-note}
**Key Finding:** COVID seasons (2020-21) show reduced home advantage across
all leagues. Post-COVID recovery varies by league.
:::

## Common Scorelines

The heatmap below shows the frequency of each scoreline (home goals x away
goals). Hover over cells to see exact percentages.

```{r vig-scoreline-heatmap}
safe_tar_read("vig_scoreline_heatmap")
```

*Scoreline frequency heatmap. X-axis: away goals (0-6+). Y-axis: home goals (0-6+). Cell colour intensity = proportion of matches with that scoreline (darker = more frequent). The most common scorelines are 1-0 (~12%), 1-1 (~11%), and 2-1 (~10%). The 0-0 scoreline accounts for ~7% of matches. Scorelines above 4-4 are extremely rare (<0.5% combined). The heatmap is asymmetric, with higher frequencies below the diagonal reflecting home advantage. Source: all parsed matches from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > FTHG / FTAG](glossary.html#data-terminology) and [Glossary > Dixon-Coles Model](glossary.html).*

::: {.callout-note}
**Key Finding:** 1-0 (~12%), 1-1 (~11%), 2-1 (~10%) are most common.
Scorelines above 4-4 are extremely rare (<0.5% combined).
:::

## Goals Per Season

Is scoring increasing over time? The chart below tracks mean total goals
per match by season and league.

```{r vig-goals-per-season-trend}
safe_tar_read("vig_goals_per_season_trend")
```

*Mean total goals per match by season and league. X-axis: season (year). Y-axis: mean total goals per match (home + away). Lines: one per league, colour-coded by league code. The Bundesliga (D1) consistently ranks among the highest-scoring leagues at approximately 3.0 goals per match. Most leagues show a slight upward trend since 2015, possibly driven by tactical evolution toward more attacking football. Ligue 1 (F1) and Serie A (I1) tend to have lower-scoring matches (~2.5 goals/match). Source: [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > FTHG / FTAG](glossary.html#data-terminology).*

::: {.callout-note}
**Key Finding:** Slight upward trend since 2015 in most leagues. Bundesliga
consistently among highest-scoring (~3.0 goals/match).
:::

## Outliers

Matches with more than 7 total goals are rare outliers. These are worth
inspecting for data quality and as edge cases for model robustness.

```{r vig-outlier-matches}
outliers <- safe_tar_read("vig_outlier_matches")
if (is.data.frame(outliers)) {
  outliers |>
    dplyr::select(-match_id) |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "High-scoring matches (>7 goals). ~0.1% of all matches."
      ),
      filter = "top",
      options = list(pageLength = 10, dom = "Bfrtip")
    )
} else {
  outliers
}
```

*High-scoring outlier matches table. Columns: league_code, season, match_date, home_team, away_team, fthg (home goals), ftag (away goals), total_goals. Filtered to matches with more than 7 total goals, representing approximately 0.1% of all matches. These extreme scorelines are important for model robustness testing --- Poisson and Dixon-Coles models may underpredict their frequency. They also serve as data quality checkpoints, since some may reflect data entry errors. Source: [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > FTHG / FTAG](glossary.html#data-terminology) and [Glossary > Poisson Regression](glossary.html).*

## Match Statistics

Median match statistics (shots, corners, fouls, cards) by league.
See [Glossary > Match Statistics](glossary.html#match-statistics) for variable definitions.

```{r vig-typical-match-stats}
match_stats <- safe_tar_read("vig_typical_match_stats")
if (is.data.frame(match_stats)) {
  match_stats |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Median match statistics by league. HS/AS = shots, HC/AC = corners, HY/AY = yellow cards."
      ),
      filter = "top",
      options = list(pageLength = 15, dom = "Bfrtip", scrollX = TRUE)
    )
} else {
  match_stats
}
```

*Median match statistics by league. Columns: league_code, HS (home shots), AS (away shots), HST (home shots on target), AST (away shots on target), HC (home corners), AC (away corners), HF (home fouls), AF (away fouls), HY (home yellow cards), AY (away yellow cards). Home teams consistently record more shots and corners than away teams across all leagues, reflecting home advantage beyond just results. La Liga (SP1) and Serie A (I1) tend to have higher foul counts and yellow cards. Bundesliga (D1) shows higher shot counts. Source: [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Match Statistics](glossary.html#match-statistics).*

## Pinnacle Overround

The [overround](https://en.wikipedia.org/wiki/Vigorish) is the bookmaker's
built-in margin. Pinnacle's margin (~2-4%) is the lowest in the industry.

```{r vig-overround-by-league}
safe_tar_read("vig_overround_by_league")
```

*Pinnacle 1X2 overround distribution by league. X-axis: league code. Y-axis: overround percentage (sum of implied probabilities minus 100%). Box plots show the distribution across matches within each league. Median overround is approximately 2-4%, far below soft bookmakers which charge 5-15%. The overround is remarkably consistent across leagues, indicating Pinnacle applies a uniform margin policy. Lower overround means the market is closer to fair odds, making Pinnacle the strongest benchmark for model evaluation. Source: Pinnacle closing odds from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Overround](glossary.html) and [Glossary > Fair Odds](glossary.html).*

::: {.callout-note}
**Key Finding:** Pinnacle margin is ~2-4%, far below soft bookmakers (5-15%).
Consistent across all leagues.
:::

## Pinnacle Calibration

A well-calibrated bookmaker's implied probabilities should match actual outcome
frequencies.

```{r vig-odds-vs-result}
safe_tar_read("vig_odds_vs_result")
```

*Pinnacle calibration plot (reliability diagram). X-axis: Pinnacle implied probability (binned). Y-axis: observed outcome frequency. Diagonal line = perfect calibration. Points near the diagonal indicate well-calibrated probabilities. Pinnacle is near-perfectly calibrated across most of the probability range. A slight favourite-longshot bias appears at the extremes: favourites (high implied probability) win slightly less often than implied, while longshots (low implied probability) win slightly more often. This near-perfect calibration makes Pinnacle the gold-standard benchmark for model comparison. Source: Pinnacle closing odds vs actual results from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Implied Probability](glossary.html) and [Glossary > Closing Line Value](glossary.html).*

::: {.callout-note}
**Key Finding:** Pinnacle is near-perfectly calibrated (points near diagonal).
Slight favourite-longshot bias at extremes.
:::

## Elo Ratings

[Elo ratings](https://en.wikipedia.org/wiki/Elo_rating_system) quantify
relative team strength, updated match-by-match with
[COOPER/FiveThirtyEight enhancements](https://www.natesilver.net/p/introducing-cooper-silver-bulletins):
dynamic K-factor, margin weighting, league reversion, and asymmetric wins.

```{r vig-elo-spread-plot}
safe_tar_read("vig_elo_spread_plot")
```

*Elo rating spread by league. X-axis: league code. Y-axis: Elo rating. Box plots show the distribution of current team Elo ratings within each league. Top-division leagues (E0, D1, I1, SP1, F1) have wider spreads, indicating greater competitive imbalance between elite and weaker teams. Division 2 leagues show more compressed distributions, reflecting more even competition. Outliers at the top represent dominant teams (e.g., Bayern Munich in D1, PSG in F1). The overall median is approximately 1500 (starting rating). Source: Elo ratings computed via `compute_elo()` on [football-data.co.uk](https://www.football-data.co.uk/) data. See [Glossary > Elo Rating](glossary.html).*

::: {.callout-note}
**Key Finding:** Top divisions have wider Elo spreads (more competitive
imbalance). Division 2 leagues show more compressed rating distributions.
:::
