---
title: "Exploratory Data Analysis"
subtitle: "Statistical patterns in European football"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    page-layout: full
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Exploratory Data Analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(htmltools)
library(DT)
library(dplyr)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) {
    rds_path <- system.file(file.path("extdata", "vignettes", paste0(name, ".rds")), package = "footbet")
    if (!nzchar(rds_path) || !file.exists(rds_path))
      rds_path <- here::here("inst", "extdata", "vignettes", paste0(name, ".rds"))
    if (file.exists(rds_path)) result <- readRDS(rds_path)
  }
  if (is.null(result)) {
    htmltools::div(style = "background: #dc3545; color: white; padding: 1em; border-radius: 4px;",
      htmltools::p(paste0("[MISSING EVIDENCE] Target `", name, "` not found.")))
  } else result
}
```

Part 3 of the [footbet](https://johngavin.github.io/footbet/) analytics guide.
See also: [Data Sources](data-sources.html) |
[Data Cleaning](data-cleaning.html) | [Models & Betting](models-betting.html) |
[Glossary](glossary.html)

This section explores key statistical patterns in European football match
data across 10 leagues and multiple seasons.

## Goal Distributions

Football goals are often modelled as
[Poisson-distributed](https://en.wikipedia.org/wiki/Poisson_distribution)
random variables. The chart below compares the observed goal distribution
against a Poisson overlay for home and away teams.

```{r vig-goals-distribution}
safe_tar_read("vig_goals_distribution")
```

```{r vig-poisson-test}
safe_tar_read("vig_poisson_test")
```

::: {.callout-note}
**Key Finding:** Home teams score slightly more than away teams (home
advantage). The Poisson distribution is a reasonable first approximation,
though it underpredicts 0-0 draws and some high-scoring outcomes.
:::

## Result Proportions

Home advantage is a well-documented phenomenon in football. The chart below
shows the proportion of Home wins, Draws, and Away wins by league, compared
to a uniform 33/33/33 baseline.

```{r vig-result-proportions}
safe_tar_read("vig_result_proportions")
```

::: {.callout-note}
**Key Finding:** Home win rate ranges 43-47% across leagues. Draws: 25-28%.
Away wins: 27-30%. All leagues show significant home advantage.
:::

## Home Advantage Trends

Home advantage has declined in recent years, with a notable drop during the
COVID-19 pandemic (2020-21 season) when matches were played in empty stadiums.

```{r vig-home-advantage-by-league}
safe_tar_read("vig_home_advantage_by_league")
```

```{r vig-home-trend-test}
safe_tar_read("vig_home_trend_test")
```

::: {.callout-note}
**Key Finding:** COVID seasons (2020-21) show reduced home advantage across
all leagues. Post-COVID recovery varies by league.
:::

## Common Scorelines

The heatmap below shows the frequency of each scoreline (home goals x away
goals). Hover over cells to see exact percentages.

```{r vig-scoreline-heatmap}
safe_tar_read("vig_scoreline_heatmap")
```

::: {.callout-note}
**Key Finding:** 1-0 (~12%), 1-1 (~11%), 2-1 (~10%) are most common.
Scorelines above 4-4 are extremely rare (<0.5% combined).
:::

## Goals Per Season

Is scoring increasing over time? The chart below tracks mean total goals
per match by season and league.

```{r vig-goals-per-season-trend}
safe_tar_read("vig_goals_per_season_trend")
```

::: {.callout-note}
**Key Finding:** Slight upward trend since 2015 in most leagues. Bundesliga
consistently among highest-scoring (~3.0 goals/match).
:::

## Outliers

Matches with more than 7 total goals are rare outliers. These are worth
inspecting for data quality and as edge cases for model robustness.

```{r vig-outlier-matches}
outliers <- safe_tar_read("vig_outlier_matches")
if (is.data.frame(outliers)) {
  outliers |>
    dplyr::select(-match_id) |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "High-scoring matches (>7 goals). ~0.1% of all matches."
      ),
      filter = "top",
      options = list(pageLength = 10, dom = "Bfrtip")
    )
} else {
  outliers
}
```

## Match Statistics

Median match statistics (shots, corners, fouls, cards) by league.
See [Glossary > Match Statistics](glossary.html#match-statistics) for variable definitions.

```{r vig-typical-match-stats}
match_stats <- safe_tar_read("vig_typical_match_stats")
if (is.data.frame(match_stats)) {
  match_stats |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Median match statistics by league. HS/AS = shots, HC/AC = corners, HY/AY = yellow cards."
      ),
      filter = "top",
      options = list(pageLength = 15, dom = "Bfrtip", scrollX = TRUE)
    )
} else {
  match_stats
}
```

## Pinnacle Overround

The [overround](https://en.wikipedia.org/wiki/Vigorish) is the bookmaker's
built-in margin. Pinnacle's margin (~2-4%) is the lowest in the industry.

```{r vig-overround-by-league}
safe_tar_read("vig_overround_by_league")
```

::: {.callout-note}
**Key Finding:** Pinnacle margin is ~2-4%, far below soft bookmakers (5-15%).
Consistent across all leagues.
:::

## Pinnacle Calibration

A well-calibrated bookmaker's implied probabilities should match actual outcome
frequencies.

```{r vig-odds-vs-result}
safe_tar_read("vig_odds_vs_result")
```

::: {.callout-note}
**Key Finding:** Pinnacle is near-perfectly calibrated (points near diagonal).
Slight favourite-longshot bias at extremes.
:::

## Elo Ratings

[Elo ratings](https://en.wikipedia.org/wiki/Elo_rating_system) quantify
relative team strength, updated match-by-match with
[COOPER/FiveThirtyEight enhancements](https://www.natesilver.net/p/introducing-cooper-silver-bulletins):
dynamic K-factor, margin weighting, league reversion, and asymmetric wins.

```{r vig-elo-spread-plot}
safe_tar_read("vig_elo_spread_plot")
```

::: {.callout-note}
**Key Finding:** Top divisions have wider Elo spreads (more competitive
imbalance). Division 2 leagues show more compressed rating distributions.
:::
