---
title: "Football Betting Analytics"
subtitle: "Data, Analysis, Models & Strategy"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    page-layout: full
    code-fold: true
    include-in-header:
      text: |
        <script src="https://cdn.jsdelivr.net/npm/mermaid/dist/mermaid.min.js"></script>
        <script>
          mermaid.initialize({
            startOnLoad: true,
            securityLevel: 'loose',
            theme: 'default'
          });
        </script>
vignette: >
  %\VignetteIndexEntry{Football Betting Analytics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE
)
library(targets)
library(htmltools)
library(DT)
library(dplyr)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  # Try targets store first
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)

  # Fallback to pre-computed RDS (for CI builds without _targets)
  if (is.null(result)) {
    # Try installed package location first
    rds_path <- system.file(
      file.path("extdata", "vignettes", paste0(name, ".rds")),
      package = "footbet"
    )

    # Fallback to source inst/ directory (for pkgdown build from source)
    if (!nzchar(rds_path) || !file.exists(rds_path)) {
      rds_path <- here::here("inst", "extdata", "vignettes", paste0(name, ".rds"))
    }

    if (file.exists(rds_path)) {
      result <- readRDS(rds_path)
    }
  }

  # Return visible error if still missing
  if (is.null(result)) {
    htmltools::div(
      style = paste(
        "background: #dc3545; color: white; padding: 1em;",
        "border-radius: 4px; margin: 1em 0; font-weight: bold;"
      ),
      htmltools::p(
        paste0("[MISSING EVIDENCE] Target `", name, "` not found.")
      ),
      htmltools::p(
        "Run `targets::tar_make()` to build vignette outputs.",
        style = "font-weight: normal; font-size: 0.9em;"
      )
    )
  } else {
    result
  }
}
```

This vignette provides a comprehensive overview of the
[footbet](https://johngavin.github.io/footbet/) package for European football
betting analytics. Navigate using the tabs below.

::: {.panel-tabset}

## Data Sources

The [footbet](https://johngavin.github.io/footbet/) package downloads match
data from [football-data.co.uk](https://www.football-data.co.uk/), a free
source of European football results and bookmaker odds maintained since the
mid-1990s. Each CSV file contains one league-season and includes full-time and
half-time scores, match statistics (shots, corners, cards), and closing odds
from multiple bookmakers including
[Pinnacle](https://en.wikipedia.org/wiki/Pinnacle_(sportsbook)) --- widely
regarded as the sharpest bookmaker in the industry.

The URL pattern is `https://www.football-data.co.uk/mmz4281/{season}/{league}.csv`,
where `{season}` is a 4-digit code (e.g., `2324` for 2023-24) and `{league}`
is a 2-character code (e.g., `E0` for the English Premier League).

`footbet::target_leagues()` returns 10 leagues covering the top 2 divisions
of the "big five" European football countries.

### Data Pipeline Overview

```{r vig-pipeline-diagram}
#| echo: false
#| output: asis
safe_tar_read("vig_pipeline_html")
```

*Data pipeline: CSV download → parsing → validation → modelling → evaluation. Nodes link to package functions.*

```{r vig-data-source-summary}
data_summary <- safe_tar_read("vig_data_source_summary")
if (is.data.frame(data_summary)) {
  data_summary |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Data coverage by league. 10 leagues across 5 countries, top 2 divisions each."
      ),
      options = list(pageLength = 15, dom = "t")
    )
} else {
  data_summary
}
```

### Coverage Grid

The grid below shows the number of parsed matches per league-season cell.
Zeroes indicate seasons where the CSV was unavailable or empty.

```{r vig-league-season-grid}
grid <- safe_tar_read("vig_league_season_grid")
if (is.data.frame(grid)) {
  grid |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Matches per league-season. 0 = data unavailable. ~380 for Div 1, ~460+ for Div 2."
      ),
      options = list(pageLength = 15, dom = "t", scrollX = TRUE)
    )
} else {
  grid
}
```

```{r vig-matches-per-season-plot}
safe_tar_read("vig_matches_per_season_plot")
```

::: {.callout-note}
**Key Finding:** Division 1 leagues (E0, D1, I1, SP1, F1) have ~380
matches/season. Division 2 leagues have more (~460-550 matches/season).
:::

### Pinnacle Odds

Not all leagues and seasons have Pinnacle odds. The table below shows the
percentage of matches with complete Pinnacle 1X2 closing odds (PSH, PSD, PSA)
and over/under odds.

Pinnacle coverage dropped significantly for some leagues from mid-2025 due to
a feed break at football-data.co.uk.

```{r vig-odds-columns-available}
odds_avail <- safe_tar_read("vig_odds_columns_available")
if (is.data.frame(odds_avail)) {
  odds_avail |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Pinnacle odds coverage. pct_1x2 = % with home/draw/away odds. pct_over_under = % with O/U 2.5."
      ),
      filter = "top",
      options = list(pageLength = 15, dom = "Bfrtip", scrollX = TRUE)
    )
} else {
  odds_avail
}
```

### Column Schema

After parsing, `footbet::parse_fd_csv()` produces a standardised tibble with
the following columns:

| Column | Description | Type |
|--------|-------------|------|
| `match_id` | Deterministic ID: `league_date_home_away` | character |
| `league_code` | 2-char league code (e.g., E0, D1) | character |
| `season` | 4-digit season code (e.g., 2324) | character |
| `match_date` | Match date | Date |
| `home_team` / `away_team` | Team names | character |
| `fthg` / `ftag` | Full-time home/away goals ([FTHG/FTAG](#fthg-ftag)) | integer |
| `ftr` | Full-time result: H, D, or A ([FTR](#ftr-full-time-result)) | character |
| `hthg` / `htag` | Half-time home/away goals ([HTHG/HTAG](#hthg-htag-htr)) | integer |
| `htr` | Half-time result ([HTR](#hthg-htag-htr)) | character |
| `hs` / `as_` | Home/away shots ([HS/AS](#match-statistics)) | integer |
| `hst` / `ast` | Home/away shots on target ([HST/AST](#match-statistics)) | integer |
| `hf` / `af` | Home/away fouls ([HF/AF](#match-statistics)) | integer |
| `hc` / `ac` | Home/away corners ([HC/AC](#match-statistics)) | integer |
| `hy` / `ay` | Home/away yellow cards ([HY/AY](#match-statistics)) | integer |
| `hr` / `ar` | Home/away red cards ([HR/AR](#match-statistics)) | integer |

: Columns produced by `parse_fd_csv()`. Cross-references link to the Glossary. {.striped}

`footbet::parse_fd_odds()` extracts Pinnacle closing odds:

| Column | Description |
|--------|-------------|
| `psh` / `psd` / `psa` | Pinnacle closing 1X2 odds (Home/Draw/Away) |
| `p_over25` / `p_under25` | Pinnacle over/under 2.5 goals odds |

: Columns produced by `parse_fd_odds()`. Pinnacle odds serve as the sharp-market benchmark. {.striped}

## Cleaning

The footbet pipeline includes automated quality control checks that run after
data parsing. This section summarises the QC outputs.

### Match Completeness

The heatmap below shows the percentage of matches per league-season with a
valid full-time result (FTR). A value below 100% indicates matches where the
result is missing --- typically from ongoing or postponed fixtures in the current
season.

```{r vig-completeness-plot}
safe_tar_read("vig_completeness_plot")
```

::: {.callout-note}
**Key Finding:** Most historical league-seasons achieve 100% completeness.
Current season shows partial data (matches not yet played).
:::

### Pinnacle Coverage

Pinnacle odds are the primary benchmark for model evaluation. The heatmap
shows the percentage of matches with all three Pinnacle 1X2 closing odds
present.

Coverage gaps arise from:

- **Division 2 leagues**: Pinnacle may not offer odds for lower divisions
- **Feed breaks**: The football-data.co.uk Pinnacle feed experienced a break
  from mid-2025
- **Early seasons**: Pinnacle data availability improves in more recent seasons

```{r vig-pinnacle-coverage-plot}
safe_tar_read("vig_pinnacle_coverage_plot")
```

::: {.callout-note}
**Key Finding:** Top divisions have near-100% coverage from ~2016 onward.
Division 2 coverage is variable. 0% coverage indicates feed breaks or no
Pinnacle market.
:::

### Missing Data

The chart below shows the percentage of missing values for each
column across all parsed matches. Match statistics (shots, corners, cards) are
more frequently missing than core fields.

```{r vig-missing-data-by-column}
safe_tar_read("vig_missing_data_heatmap")
```

The table below provides detailed missing counts with variable descriptions.
See [Glossary > Data Terminology](#data-terminology) for full definitions.

```{r vig-missing-data-table}
missing_data <- safe_tar_read("vig_missing_data_by_column")
if (is.data.frame(missing_data)) {
  missing_data |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Missing data by column. Match statistics often unavailable in older seasons."
      ),
      filter = "top",
      options = list(pageLength = 15, dom = "Bfrtip")
    )
} else {
  missing_data  # Show error div if target missing
}
```

### Anomaly Detection

The pipeline flags matches with:

- **Extreme scorelines**: Total goals > 10 (very rare, may indicate data
  entry errors)
- **Missing dates**: Matches without a parseable date
- **Future matches**: Dates more than 7 days beyond the pipeline run date
- **Missing team names**: Rows with NA home or away team

```{r vig-anomalies-table}
anomalies <- safe_tar_read("vig_anomalies_table")
if (is.data.frame(anomalies) && nrow(anomalies) > 0) {
  anomalies |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Flagged anomalies for manual review. Empty table = no anomalies detected."
      ),
      filter = "top",
      options = list(pageLength = 10, dom = "Bfrtip")
    )
} else if (is.data.frame(anomalies)) {
  htmltools::div(
    style = "background: #28a745; color: white; padding: 1em; border-radius: 4px;",
    "No anomalies detected in the dataset."
  )
} else {
  anomalies
}
```

### QC Summary

The `qc_summary` target combines all QC checks into a single report with a
timestamp. The pipeline computes:

- **Total matches parsed**: Row count across all leagues/seasons
- **Total odds rows**: Pinnacle odds records parsed
- **Completeness**: Percentage of matches with valid results
- **Pinnacle coverage**: Percentage with 1X2 odds
- **Anomaly count**: Flagged rows for manual review

The QC targets serve as data contracts --- downstream model targets depend on
`qc_summary` to ensure data quality issues are surfaced before model fitting.

## EDA

This section explores key statistical patterns in European football match
data across 10 leagues and multiple seasons. All visualizations are interactive
with dark theme styling.

### Goal Distributions

Football goals are often modelled as
[Poisson-distributed](https://en.wikipedia.org/wiki/Poisson_distribution)
random variables. The chart below compares the observed goal distribution
against a Poisson overlay for home and away teams.

```{r vig-goals-distribution}
safe_tar_read("vig_goals_distribution")
```

```{r vig-poisson-test}
safe_tar_read("vig_poisson_test")
```

::: {.callout-note}
**Key Finding:** Home teams score slightly more than away teams (home
advantage). The Poisson distribution is a reasonable first approximation,
though it underpredicts 0-0 draws and some high-scoring outcomes. See Dixon &
Coles (1997) for the correlation correction term.
:::

### Result Proportions

Home advantage is a well-documented phenomenon in football. The chart below
shows the proportion of Home wins, Draws, and Away wins by league, compared
to a uniform 33/33/33 baseline.

```{r vig-result-proportions}
safe_tar_read("vig_result_proportions")
```

::: {.callout-note}
**Key Finding:** Home win rate ranges 43-47% across leagues. Draws: 25-28%.
Away wins: 27-30%. All leagues show significant home advantage.
:::

### Home Advantage Trends

Home advantage has declined in recent years, with a notable drop during the
COVID-19 pandemic (2020-21 season) when matches were played in empty stadiums.

```{r vig-home-advantage-by-league}
safe_tar_read("vig_home_advantage_by_league")
```

```{r vig-home-trend-test}
safe_tar_read("vig_home_trend_test")
```

::: {.callout-note}
**Key Finding:** COVID seasons (2020-21) show reduced home advantage across
all leagues. Post-COVID recovery varies by league. Long-term declining trend
visible in most leagues.
:::

### Common Scorelines

The heatmap below shows the frequency of each scoreline (home goals x away
goals). Hover over cells to see exact percentages.

```{r vig-scoreline-heatmap}
safe_tar_read("vig_scoreline_heatmap")
```

::: {.callout-note}
**Key Finding:** 1-0 (~12%), 1-1 (~11%), 2-1 (~10%) are most common.
Scorelines above 4-4 are extremely rare (<0.5% combined).
:::

### Goals Per Season

Is scoring increasing over time? The chart below tracks mean total goals
per match by season and league. Use the range slider to zoom into specific periods.

```{r vig-goals-per-season-trend}
safe_tar_read("vig_goals_per_season_trend")
```

::: {.callout-note}
**Key Finding:** Slight upward trend since 2015 in most leagues. Bundesliga
consistently among highest-scoring (~3.0 goals/match). Serie A shows notable
increase in recent seasons.
:::

### Outliers

Matches with more than 7 total goals are rare outliers. These are worth
inspecting for data quality and as edge cases for model robustness.

```{r vig-outlier-matches}
outliers <- safe_tar_read("vig_outlier_matches")
if (is.data.frame(outliers)) {
  outliers |>
    dplyr::select(-match_id) |>  # Drop internal ID
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "High-scoring matches (>7 goals). ~0.1% of all matches."
      ),
      filter = "top",
      options = list(pageLength = 10, dom = "Bfrtip")
    )
} else {
  outliers  # Show error div if target missing
}
```

### Match Statistics

Median match statistics (shots, corners, fouls, cards) by league. These vary
across leagues reflecting different playing styles.
See [Glossary > Match Statistics](glossary.html#match-statistics) for variable definitions.

```{r vig-typical-match-stats}
match_stats <- safe_tar_read("vig_typical_match_stats")
if (is.data.frame(match_stats)) {
  match_stats |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Median match statistics by league. HS/AS = shots, HC/AC = corners, HY/AY = yellow cards."
      ),
      filter = "top",
      options = list(pageLength = 15, dom = "Bfrtip", scrollX = TRUE)
    )
} else {
  match_stats
}
```

### Pinnacle Overround

The [overround](https://en.wikipedia.org/wiki/Vigorish) (or vig) is the
bookmaker's built-in margin. Pinnacle's margin (~2-4%) is the lowest in the
industry, making their odds the best available benchmark for "true"
probabilities.

```{r vig-overround-by-league}
safe_tar_read("vig_overround_by_league")
```

::: {.callout-note}
**Key Finding:** Pinnacle margin is ~2-4%, far below soft bookmakers (5-15%).
Lower margin = fairer odds = better model benchmark. Consistent across all
leagues.
:::

### Pinnacle Calibration

A well-calibrated bookmaker's implied probabilities should match actual outcome
frequencies. The plot below bins devigged Pinnacle probabilities and compares
them to observed outcome rates.

```{r vig-odds-vs-result}
safe_tar_read("vig_odds_vs_result")
```

::: {.callout-note}
**Key Finding:** Pinnacle is near-perfectly calibrated (points near diagonal).
Slight favourite-longshot bias at extremes. Size indicates sample size per bin.
:::

### Elo Ratings

[Elo ratings](https://en.wikipedia.org/wiki/Elo_rating_system) quantify
relative team strength. The chart below shows the distribution of final Elo
ratings per league.

```{r vig-elo-spread-plot}
safe_tar_read("vig_elo_spread_plot")
```

::: {.callout-note}
**Key Finding:** Top divisions have wider Elo spreads (more competitive
imbalance). Division 2 leagues show more compressed rating distributions.
Initial rating = 1500 for all teams.
:::

## Models

This section covers the modelling pipeline: walk-forward cross-validation,
two statistical models, and benchmarking against Pinnacle implied probabilities.

### Cross-Validation

Football data is time-ordered, so standard k-fold CV would leak future
information. The pipeline uses **walk-forward** (expanding window) splits:

- **Training window**: 24 months of historical matches
- **Test window**: 1 month of subsequent matches
- The window slides forward by 1 month for each fold

This mimics a realistic betting scenario where you train on past data and
predict the next month.

`footbet::walk_forward_splits()` generates the date-based train/test splits,
and `evaluate_glm_baseline()` / `evaluate_dc()` iterate over them.

```{r vig-cv-diagram}
#| echo: false
#| output: asis
safe_tar_read("vig_cv_html")
```

*Walk-forward cross-validation: 24-month rolling training window, 1-month test.*

### GLM Poisson

The baseline model is a
[Poisson regression](https://en.wikipedia.org/wiki/Poisson_regression)
for goal counts, following Maher (1982):

$$\text{goals}_{ij} \sim \text{Poisson}(\lambda_{ij})$$

$$\log(\lambda_{ij}) = \mu + \alpha_i + \beta_j + \gamma \cdot \mathbb{1}[\text{home}]$$

where $\alpha_i$ is team $i$'s attack strength, $\beta_j$ is team $j$'s
defence weakness, and $\gamma$ captures home advantage. The model assumes
independence between home and away goals.

`footbet::fit_poisson_glm()` fits this model; `predict_glm()` generates a
bivariate score probability matrix (up to 10x10 goals) and derives 1X2,
over/under, and Asian handicap probabilities.

### Dixon-Coles

Dixon & Coles (1997) extend the Poisson model with:

1. **Correlation correction**: A parameter $\rho$ adjusts the joint
   probability of low-scoring outcomes (0-0, 1-0, 0-1, 1-1) where the
   independence assumption fails most
2. **Time-decay weights**: Recent matches receive higher weight via an
   exponential decay parameter $\xi$

The pipeline uses $\xi = 0.003$ (approximately 1-year half-life) via the
[goalmodel](https://cran.r-project.org/package=goalmodel) package.

### CV Metrics

The plot below shows
[log-loss](https://en.wikipedia.org/wiki/Cross-entropy),
[Brier score](https://en.wikipedia.org/wiki/Brier_score), and
[RPS](https://en.wikipedia.org/wiki/Ranked_probability_score)
per CV fold for both models, with Pinnacle as a benchmark.
All three are **proper scoring rules** --- lower is better.

```{r vig-cv-metrics-plot}
safe_tar_read("vig_cv_metrics_plot")
```

```{r vig-model-comparison-test}
safe_tar_read("vig_model_comparison_test")
```

::: {.callout-note}
**Key Finding:** Dixon-Coles slightly outperforms GLM. Both models trail
Pinnacle on most folds. Pinnacle's edge is consistent across metrics.
:::

### Model Comparison

The table below summarises mean metrics across all folds, with the Pinnacle
benchmark for comparison. Positive "edge" means the model outperforms Pinnacle.

```{r vig-model-comparison-table}
model_comp <- safe_tar_read("vig_model_comparison_table")
if (is.data.frame(model_comp)) {
  model_comp |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Model comparison. Lower scores = better. Positive edge = outperforms Pinnacle."
      ),
      options = list(pageLength = 10, dom = "t")
    )
} else {
  model_comp
}
```

### Value Bets

A **value bet** exists when the model's estimated probability exceeds the
market's implied probability by a minimum edge threshold. The pipeline uses:

- **Minimum edge**: 3% (model_prob - market_prob > 0.03)
- **Odds filter**: Only bets between 1.50 and 10.00 decimal odds
- **Devigged odds**: Market probabilities via [Shin (1993)](https://doi.org/10.2307/2234526) method

```{r vig-value-bets-summary}
vb_summary <- safe_tar_read("vig_value_bets_summary")
if (is.data.frame(vb_summary)) {
  vb_summary |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Value bet summary by outcome and league. Edge = model prob - market prob."
      ),
      filter = "top",
      options = list(pageLength = 15, dom = "Bfrtip")
    )
} else {
  vb_summary
}
```

```{r vig-edge-distribution}
safe_tar_read("vig_edge_distribution")
```

::: {.callout-note}
**Key Finding:** Most edges cluster at 3-8%. Large edges (>15%) are rare and
may signal model error.
:::

### Kelly Staking

The pipeline uses **quarter Kelly** staking:

$$f^* = \frac{1}{4} \cdot \frac{p \cdot (b + 1) - 1}{b}$$

where $p$ is the model probability and $b$ is the decimal odds minus 1.

**Guardrails:**

- **Max stake**: 3% of current bankroll per bet
- **Drawdown halt**: Stakes halved when bankroll drops 20% from peak
- **`apply_guardrails()`**: Automatically reduces stake size during drawdowns

```{r vig-kelly-diagram}
#| echo: false
#| output: asis
safe_tar_read("vig_kelly_html")
```

*Kelly staking decision tree: edge ≥ 3% and odds 1.50–10.00 triggers quarter-Kelly stake.*

```{r vig-kelly-stake-distribution}
safe_tar_read("vig_kelly_stake_distribution")
```

### P&L Simulation

The bankroll curve below shows the simulated profit & loss trajectory for all
GLM value bets, staked with quarter Kelly and drawdown guardrails.
Use the range slider to zoom into specific periods.

```{r vig-pnl-curve}
safe_tar_read("vig_pnl_curve")
```

```{r vig-drawdown-plot}
safe_tar_read("vig_drawdown_plot")
```

::: {.callout-note}
**Key Finding:** Trajectory depends on model accuracy and market conditions.
Drawdown guardrails limit losses during poor runs. Maximum drawdown is the key
risk metric.
:::

#### Summary

```{r vig-pnl-summary-table}
pnl_summary <- safe_tar_read("vig_pnl_summary_table")
if (is.data.frame(pnl_summary)) {
  pnl_summary |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "P&L summary. ROI = total profit / total staked. Sharpe = risk-adjusted return."
      ),
      options = list(pageLength = 10, dom = "t")
    )
} else {
  pnl_summary
}
```

## Glossary

See the full [Glossary](glossary.html) for definitions of all betting,
statistical, and data terminology used in this article.
