---
title: "Football Betting Analytics"
subtitle: "Data, Analysis, Models & Strategy"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: false
    page-layout: full
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Football Betting Analytics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE
)
library(targets)
library(htmltools)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  # Try targets store first
 result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)

 # Fallback to pre-computed RDS (for CI builds without _targets)
 if (is.null(result)) {
   rds_path <- system.file(
     file.path("extdata", "vignettes", paste0(name, ".rds")),
     package = "footbet"
   )
   if (nzchar(rds_path) && file.exists(rds_path)) {
     result <- readRDS(rds_path)
   }
 }

 # Return visible error if still missing
 if (is.null(result)) {
   htmltools::div(
     style = paste(
       "background: #dc3545; color: white; padding: 1em;",
       "border-radius: 4px; margin: 1em 0; font-weight: bold;"
     ),
     htmltools::p(
       paste0("[MISSING EVIDENCE] Target `", name, "` not found.")
     ),
     htmltools::p(
       "Run `targets::tar_make()` to build vignette outputs.",
       style = "font-weight: normal; font-size: 0.9em;"
     )
   )
 } else {
   result
 }
}
```

This vignette provides a comprehensive overview of the
[footbet](https://johngavin.github.io/footbet/) package for European football
betting analytics. Navigate using the tabs below.

::: {.panel-tabset}

## Data Sources

The [footbet](https://johngavin.github.io/footbet/) package downloads match
data from [football-data.co.uk](https://www.football-data.co.uk/), a free
source of European football results and bookmaker odds maintained since the
mid-1990s. Each CSV file contains one league-season and includes full-time and
half-time scores, match statistics (shots, corners, cards), and closing odds
from multiple bookmakers including
[Pinnacle](https://en.wikipedia.org/wiki/Pinnacle_(sportsbook)) --- widely
regarded as the sharpest bookmaker in the industry.

The URL pattern is `https://www.football-data.co.uk/mmz4281/{season}/{league}.csv`,
where `{season}` is a 4-digit code (e.g., `2324` for 2023-24) and `{league}`
is a 2-character code (e.g., `E0` for the English Premier League).

`footbet::target_leagues()` returns 10 leagues covering the top 2 divisions
of the "big five" European football countries.

```{r vig-data-source-summary}
safe_tar_read("vig_data_source_summary")
```

### Coverage Grid

The grid below shows the number of parsed matches per league-season cell.
Zeroes indicate seasons where the CSV was unavailable or empty.

```{r vig-league-season-grid}
safe_tar_read("vig_league_season_grid")
```

```{r vig-matches-per-season-plot}
safe_tar_read("vig_matches_per_season_plot")
```

::: {.callout-note}
**Key Finding:** Division 1 leagues (E0, D1, I1, SP1, F1) have ~380
matches/season. Division 2 leagues have more (~460-550 matches/season).
:::

### Pinnacle Odds

Not all leagues and seasons have Pinnacle odds. The table below shows the
percentage of matches with complete Pinnacle 1X2 closing odds (PSH, PSD, PSA)
and over/under odds.

Pinnacle coverage dropped significantly for some leagues from mid-2025 due to
a feed break at football-data.co.uk.

```{r vig-odds-columns-available}
safe_tar_read("vig_odds_columns_available")
```

### Column Schema

After parsing, `footbet::parse_fd_csv()` produces a standardised tibble with
the following columns:

| Column | Description | Type |
|--------|-------------|------|
| `match_id` | Deterministic ID: `league_date_home_away` | character |
| `league_code` | 2-char league code (e.g., E0, D1) | character |
| `season` | 4-digit season code (e.g., 2324) | character |
| `match_date` | Match date | Date |
| `home_team` / `away_team` | Team names | character |
| `fthg` / `ftag` | Full-time home/away goals | integer |
| `ftr` | Full-time result: H, D, or A | character |
| `hthg` / `htag` | Half-time home/away goals | integer |
| `htr` | Half-time result | character |
| `hs` / `as_` | Home/away shots | integer |
| `hst` / `ast` | Home/away shots on target | integer |
| `hf` / `af` | Home/away fouls | integer |
| `hc` / `ac` | Home/away corners | integer |
| `hy` / `ay` | Home/away yellow cards | integer |
| `hr` / `ar` | Home/away red cards | integer |

`footbet::parse_fd_odds()` extracts Pinnacle closing odds:

| Column | Description |
|--------|-------------|
| `psh` / `psd` / `psa` | Pinnacle closing 1X2 odds (Home/Draw/Away) |
| `p_over25` / `p_under25` | Pinnacle over/under 2.5 goals odds |

## Cleaning

The footbet pipeline includes automated quality control checks that run after
data parsing. This section summarises the QC outputs.

### Match Completeness

The heatmap below shows the percentage of matches per league-season with a
valid full-time result (FTR). A value below 100% indicates matches where the
result is missing --- typically from ongoing or postponed fixtures in the current
season.

```{r vig-completeness-plot}
safe_tar_read("vig_completeness_plot")
```

::: {.callout-note}
**Key Finding:** Most historical league-seasons achieve 100% completeness.
Current season shows partial data (matches not yet played).
:::

### Pinnacle Coverage

Pinnacle odds are the primary benchmark for model evaluation. The heatmap
shows the percentage of matches with all three Pinnacle 1X2 closing odds
present.

Coverage gaps arise from:

- **Division 2 leagues**: Pinnacle may not offer odds for lower divisions
- **Feed breaks**: The football-data.co.uk Pinnacle feed experienced a break
  from mid-2025
- **Early seasons**: Pinnacle data availability improves in more recent seasons

```{r vig-pinnacle-coverage-plot}
safe_tar_read("vig_pinnacle_coverage_plot")
```

::: {.callout-note}
**Key Finding:** Top divisions have near-100% coverage from ~2016 onward.
Division 2 coverage is variable. 0% coverage indicates feed breaks or no
Pinnacle market.
:::

### Missing Data

The chart below shows the percentage of missing values for each
column across all parsed matches. Match statistics (shots, corners, cards) are
more frequently missing than core fields.

```{r vig-missing-data-by-column}
safe_tar_read("vig_missing_data_heatmap")
```

The table below provides detailed missing counts:

```{r vig-missing-data-table}
safe_tar_read("vig_missing_data_by_column")
```

### Anomaly Detection

The pipeline flags matches with:

- **Extreme scorelines**: Total goals > 10 (very rare, may indicate data
  entry errors)
- **Missing dates**: Matches without a parseable date
- **Future matches**: Dates more than 7 days beyond the pipeline run date
- **Missing team names**: Rows with NA home or away team

```{r vig-anomalies-table}
safe_tar_read("vig_anomalies_table")
```

### QC Summary

The `qc_summary` target combines all QC checks into a single report with a
timestamp. The pipeline computes:

- **Total matches parsed**: Row count across all leagues/seasons
- **Total odds rows**: Pinnacle odds records parsed
- **Completeness**: Percentage of matches with valid results
- **Pinnacle coverage**: Percentage with 1X2 odds
- **Anomaly count**: Flagged rows for manual review

The QC targets serve as data contracts --- downstream model targets depend on
`qc_summary` to ensure data quality issues are surfaced before model fitting.

## EDA

This section explores key statistical patterns in European football match
data across 10 leagues and multiple seasons. All visualizations are interactive
with dark theme styling.

### Goal Distributions

Football goals are often modelled as
[Poisson-distributed](https://en.wikipedia.org/wiki/Poisson_distribution)
random variables. The chart below compares the observed goal distribution
against a Poisson overlay for home and away teams.

```{r vig-goals-distribution}
safe_tar_read("vig_goals_distribution")
```

```{r vig-poisson-test}
safe_tar_read("vig_poisson_test")
```

::: {.callout-note}
**Key Finding:** Home teams score slightly more than away teams (home
advantage). The Poisson distribution is a reasonable first approximation,
though it underpredicts 0-0 draws and some high-scoring outcomes. See Dixon &
Coles (1997) for the correlation correction term.
:::

### Result Proportions

Home advantage is a well-documented phenomenon in football. The chart below
shows the proportion of Home wins, Draws, and Away wins by league, compared
to a uniform 33/33/33 baseline.

```{r vig-result-proportions}
safe_tar_read("vig_result_proportions")
```

::: {.callout-note}
**Key Finding:** Home win rate ranges 43-47% across leagues. Draws: 25-28%.
Away wins: 27-30%. All leagues show significant home advantage.
:::

### Home Advantage Trends

Home advantage has declined in recent years, with a notable drop during the
COVID-19 pandemic (2020-21 season) when matches were played in empty stadiums.

```{r vig-home-advantage-by-league}
safe_tar_read("vig_home_advantage_by_league")
```

```{r vig-home-trend-test}
safe_tar_read("vig_home_trend_test")
```

::: {.callout-note}
**Key Finding:** COVID seasons (2020-21) show reduced home advantage across
all leagues. Post-COVID recovery varies by league. Long-term declining trend
visible in most leagues.
:::

### Common Scorelines

The heatmap below shows the frequency of each scoreline (home goals x away
goals). Hover over cells to see exact percentages.

```{r vig-scoreline-heatmap}
safe_tar_read("vig_scoreline_heatmap")
```

::: {.callout-note}
**Key Finding:** 1-0 (~12%), 1-1 (~11%), 2-1 (~10%) are most common.
Scorelines above 4-4 are extremely rare (<0.5% combined).
:::

### Goals Per Season

Is scoring increasing over time? The chart below tracks mean total goals
per match by season and league. Use the range slider to zoom into specific periods.

```{r vig-goals-per-season-trend}
safe_tar_read("vig_goals_per_season_trend")
```

::: {.callout-note}
**Key Finding:** Slight upward trend since 2015 in most leagues. Bundesliga
consistently among highest-scoring (~3.0 goals/match). Serie A shows notable
increase in recent seasons.
:::

### Outliers

Matches with more than 7 total goals are rare outliers. These are worth
inspecting for data quality and as edge cases for model robustness.

```{r vig-outlier-matches}
safe_tar_read("vig_outlier_matches")
```

### Match Statistics

Median match statistics (shots, corners, fouls, cards) by league. These vary
across leagues reflecting different playing styles.

```{r vig-typical-match-stats}
safe_tar_read("vig_typical_match_stats")
```

### Pinnacle Overround

The [overround](https://en.wikipedia.org/wiki/Vigorish) (or vig) is the
bookmaker's built-in margin. Pinnacle's margin (~2-4%) is the lowest in the
industry, making their odds the best available benchmark for "true"
probabilities.

```{r vig-overround-by-league}
safe_tar_read("vig_overround_by_league")
```

::: {.callout-note}
**Key Finding:** Pinnacle margin is ~2-4%, far below soft bookmakers (5-15%).
Lower margin = fairer odds = better model benchmark. Consistent across all
leagues.
:::

### Pinnacle Calibration

A well-calibrated bookmaker's implied probabilities should match actual outcome
frequencies. The plot below bins devigged Pinnacle probabilities and compares
them to observed outcome rates.

```{r vig-odds-vs-result}
safe_tar_read("vig_odds_vs_result")
```

::: {.callout-note}
**Key Finding:** Pinnacle is near-perfectly calibrated (points near diagonal).
Slight favourite-longshot bias at extremes. Size indicates sample size per bin.
:::

### Elo Ratings

[Elo ratings](https://en.wikipedia.org/wiki/Elo_rating_system) quantify
relative team strength. The chart below shows the distribution of final Elo
ratings per league.

```{r vig-elo-spread-plot}
safe_tar_read("vig_elo_spread_plot")
```

::: {.callout-note}
**Key Finding:** Top divisions have wider Elo spreads (more competitive
imbalance). Division 2 leagues show more compressed rating distributions.
Initial rating = 1500 for all teams.
:::

## Models

This section covers the modelling pipeline: walk-forward cross-validation,
two statistical models, and benchmarking against Pinnacle implied probabilities.

### Cross-Validation

Football data is time-ordered, so standard k-fold CV would leak future
information. The pipeline uses **walk-forward** (expanding window) splits:

- **Training window**: 24 months of historical matches
- **Test window**: 1 month of subsequent matches
- The window slides forward by 1 month for each fold

This mimics a realistic betting scenario where you train on past data and
predict the next month.

`footbet::walk_forward_splits()` generates the date-based train/test splits,
and `evaluate_glm_baseline()` / `evaluate_dc()` iterate over them.

### GLM Poisson

The baseline model is a
[Poisson regression](https://en.wikipedia.org/wiki/Poisson_regression)
for goal counts, following Maher (1982):

$$\text{goals}_{ij} \sim \text{Poisson}(\lambda_{ij})$$

$$\log(\lambda_{ij}) = \mu + \alpha_i + \beta_j + \gamma \cdot \mathbb{1}[\text{home}]$$

where $\alpha_i$ is team $i$'s attack strength, $\beta_j$ is team $j$'s
defence weakness, and $\gamma$ captures home advantage. The model assumes
independence between home and away goals.

`footbet::fit_poisson_glm()` fits this model; `predict_glm()` generates a
bivariate score probability matrix (up to 10x10 goals) and derives 1X2,
over/under, and Asian handicap probabilities.

### Dixon-Coles

Dixon & Coles (1997) extend the Poisson model with:

1. **Correlation correction**: A parameter $\rho$ adjusts the joint
   probability of low-scoring outcomes (0-0, 1-0, 0-1, 1-1) where the
   independence assumption fails most
2. **Time-decay weights**: Recent matches receive higher weight via an
   exponential decay parameter $\xi$

The pipeline uses $\xi = 0.003$ (approximately 1-year half-life) via the
[goalmodel](https://cran.r-project.org/package=goalmodel) package.

### CV Metrics

The plot below shows
[log-loss](https://en.wikipedia.org/wiki/Cross-entropy),
[Brier score](https://en.wikipedia.org/wiki/Brier_score), and
[RPS](https://en.wikipedia.org/wiki/Ranked_probability_score)
per CV fold for both models, with Pinnacle as a benchmark.
All three are **proper scoring rules** --- lower is better.

```{r vig-cv-metrics-plot}
safe_tar_read("vig_cv_metrics_plot")
```

```{r vig-model-comparison-test}
safe_tar_read("vig_model_comparison_test")
```

::: {.callout-note}
**Key Finding:** Dixon-Coles slightly outperforms GLM. Both models trail
Pinnacle on most folds. Pinnacle's edge is consistent across metrics.
:::

### Model Comparison

The table below summarises mean metrics across all folds, with the Pinnacle
benchmark for comparison. Positive "edge" means the model outperforms Pinnacle.

```{r vig-model-comparison-table}
safe_tar_read("vig_model_comparison_table")
```

### Value Bets

A **value bet** exists when the model's estimated probability exceeds the
market's implied probability by a minimum edge threshold. The pipeline uses:

- **Minimum edge**: 3% (model_prob - market_prob > 0.03)
- **Odds filter**: Only bets between 1.50 and 10.00 decimal odds
- **Devigged odds**: Market probabilities via [Shin (1993)](https://doi.org/10.2307/2234526) method

```{r vig-value-bets-summary}
safe_tar_read("vig_value_bets_summary")
```

```{r vig-edge-distribution}
safe_tar_read("vig_edge_distribution")
```

::: {.callout-note}
**Key Finding:** Most edges cluster at 3-8%. Large edges (>15%) are rare and
may signal model error.
:::

### Kelly Staking

The pipeline uses **quarter Kelly** staking:

$$f^* = \frac{1}{4} \cdot \frac{p \cdot (b + 1) - 1}{b}$$

where $p$ is the model probability and $b$ is the decimal odds minus 1.

**Guardrails:**

- **Max stake**: 3% of current bankroll per bet
- **Drawdown halt**: Stakes halved when bankroll drops 20% from peak
- **`apply_guardrails()`**: Automatically reduces stake size during drawdowns

```{r vig-kelly-stake-distribution}
safe_tar_read("vig_kelly_stake_distribution")
```

### P&L Simulation

The bankroll curve below shows the simulated profit & loss trajectory for all
GLM value bets, staked with quarter Kelly and drawdown guardrails.
Use the range slider to zoom into specific periods.

```{r vig-pnl-curve}
safe_tar_read("vig_pnl_curve")
```

```{r vig-drawdown-plot}
safe_tar_read("vig_drawdown_plot")
```

::: {.callout-note}
**Key Finding:** Trajectory depends on model accuracy and market conditions.
Drawdown guardrails limit losses during poor runs. Maximum drawdown is the key
risk metric.
:::

#### Summary

```{r vig-pnl-summary-table}
safe_tar_read("vig_pnl_summary_table")
```

## Glossary

This glossary defines betting and statistical terms used throughout the
[footbet](https://johngavin.github.io/footbet/) package.

### Odds Formats

#### Decimal Odds

The total payout per unit staked, including the original stake. A decimal
odds of 2.50 means a 1 unit stake returns 2.50 units if the bet wins
(1 unit profit + 1 unit stake). Standard in Europe and Australia.

**Example:** Decimal 2.50 on a 10 unit stake pays 25 units (15 profit).

#### Fractional Odds

Expresses profit relative to stake as a fraction. Traditional UK format.
"3/2" (read "three to two") means 3 units profit for every 2 units staked.

**Conversion:** Decimal = (numerator / denominator) + 1

**Example:** 3/2 fractional = 2.50 decimal

#### American Odds

Positive numbers show profit on a 100 unit stake; negative numbers show
stake needed to win 100 units. Standard in the United States.

- +150 means bet 100 to win 150 profit
- -200 means bet 200 to win 100 profit

**Conversion:**

- Positive: Decimal = (American / 100) + 1
- Negative: Decimal = (100 / |American|) + 1

### Market Types

#### 1X2 (Match Result)

The most common football betting market:

- **1 (H)**: Home win
- **X (D)**: Draw
- **2 (A)**: Away win

#### Over/Under (O/U)

Total goals in the match relative to a line, usually 2.5. "Over 2.5"
requires 3+ goals; "Under 2.5" requires 0-2 goals.

#### Asian Handicap (AH)

Handicap applied to one team to eliminate the draw outcome. A -1.5 Asian
handicap on the home team means they must win by 2+ goals. Half-goal
handicaps ensure no push (void bet).

#### BTTS (Both Teams to Score)

Whether both teams score at least one goal. Binary yes/no market.

### Probability Concepts

#### Implied Probability

The probability implied by decimal odds: `1 / odds`. For odds of 2.50,
the implied probability is 40%. Markets include overround, so implied
probabilities sum to >100%.

#### Overround (Vig/Juice)

The bookmaker's margin built into odds. Calculated as the sum of implied
probabilities minus 100%. A 106% book has 6% overround.

**Example:** Home 2.00, Draw 3.50, Away 4.00

- Implied: 50% + 28.6% + 25% = 103.6%
- Overround: 3.6%

#### Fair Odds

Odds after removing the overround, representing true probabilities.
Several devigging methods exist:

- **Multiplicative:** Divide each probability by the overround
- **Power:** Raise probabilities to a power that normalizes them
- **Shin:** Models overround as protection against informed bettors

#### Closing Line Value (CLV)

The difference between your bet odds and the closing odds. Positive CLV
indicates you beat the market. Tracked via `closing_line_value()`.

### Performance Metrics

#### Log Loss

Measures prediction quality for probabilistic forecasts. Lower is better.
Heavily penalizes confident wrong predictions.

$$\text{Log Loss} = -\frac{1}{n} \sum \log(p_i)$$

where $p_i$ is the predicted probability of the actual outcome.

#### Brier Score

Mean squared error for probability predictions. Ranges 0-2 for 1X2 markets.
Lower is better. Less harsh than log loss on confident errors.

$$\text{Brier} = \frac{1}{n} \sum (p_i - o_i)^2$$

#### Ranked Probability Score (RPS)

Extension of Brier score for ordered outcomes. Better suited for 1X2
because it considers that H > D > A forms an ordering by goal difference.

#### ROI (Return on Investment)

Total profit divided by total staked, expressed as percentage.
+5% ROI means 5% profit on turnover.

#### Sharpe Ratio

Risk-adjusted return metric. Higher values indicate better returns
relative to volatility. Computed via `betting_sharpe_ratio()`.

$$\text{Sharpe} = \frac{\text{Mean Return} - r_f}{\text{Std Dev Return}} \times \sqrt{n}$$

where $r_f$ is the risk-free rate and $n$ is periods per year.

#### Maximum Drawdown

Largest peak-to-trough decline in cumulative returns. Measures worst-case
scenario during a betting sequence.

### Statistical Models

#### Poisson Regression

Assumes goal counts follow Poisson distributions with team-specific
attack and defence parameters. Foundation of most football prediction models.

#### Dixon-Coles Model

Extension of Poisson that corrects for under-prediction of 0-0, 1-0, 0-1,
and 1-1 scorelines. Adds a "rho" correlation parameter. Implemented in
`fit_dixon_coles()`.

#### Expected Goals (xG)

Probability that a shot results in a goal, based on shot characteristics
(location, body part, game state). Summed across shots to get match xG.
Better predictor of future performance than actual goals.

#### Elo Rating

Rating system that updates after each match based on result versus
expectation. Higher-rated teams are expected to beat lower-rated teams.
Implemented in `compute_elo()`.

### Betting Strategy

#### Kelly Criterion

Optimal bet sizing formula that maximizes log wealth growth:

$$f^* = \frac{p \cdot b - q}{b}$$

where $p$ is win probability, $q = 1-p$, and $b$ is net odds (decimal - 1).

#### Value Bet

A bet where your estimated probability exceeds the implied probability.
If you estimate 50% win chance and odds imply 40%, that's a value bet.

#### Edge

The expected profit per unit staked. For a value bet:

$$\text{Edge} = p \times (o - 1) - (1 - p)$$

where $p$ is true probability and $o$ is decimal odds.

#### Sharp Bookmaker

Bookmaker with accurate odds that quickly incorporate new information.
Pinnacle is considered the sharpest. Their closing line is the best
publicly available proxy for true probabilities.

#### Soft Bookmaker

Bookmaker with less accurate odds, slower to adjust. May offer temporary
value but often limits winning accounts.

### Data Terminology

#### FTR (Full Time Result)

Match outcome: H (home win), D (draw), A (away win).

#### FTHG / FTAG

Full Time Home Goals / Full Time Away Goals.

#### HTHG / HTAG / HTR

Half Time Home Goals / Half Time Away Goals / Half Time Result.

#### Match Statistics

Common abbreviations in football-data.co.uk files:

- **HS/AS**: Home/Away Shots
- **HST/AST**: Home/Away Shots on Target
- **HC/AC**: Home/Away Corners
- **HF/AF**: Home/Away Fouls
- **HY/AY**: Home/Away Yellow Cards
- **HR/AR**: Home/Away Red Cards

### See Also

- [Data Sources tab](#data-sources) for data download and parsing
- [Models tab](#models) for model implementation details
- Package help: `?closing_line_value`, `?betting_sharpe_ratio`, `?convert_odds`

:::
