---
title: "Glossary"
format:
  html:
    theme: darkly
    code-fold: true
    code-summary: "Show code"
    include-in-header:
      text: |
        <script>document.documentElement.setAttribute('data-bs-theme','dark');</script>
        <style>html,body{background:#000!important;color:#fff!important;}</style>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) {
    rds_path <- system.file(file.path("extdata", "vignettes", paste0(name, ".rds")), package = "footbet")
    if (!nzchar(rds_path) || !file.exists(rds_path))
      rds_path <- here::here("inst", "extdata", "vignettes", paste0(name, ".rds"))
    if (file.exists(rds_path)) result <- readRDS(rds_path)
  }
  if (is.null(result)) {
    htmltools::div(style = "background: #dc3545; color: white; padding: 1em; border-radius: 4px;",
      htmltools::p(paste0("[MISSING EVIDENCE] Target `", name, "` not found.")))
  } else result
}
```

This glossary defines betting and statistical terms used throughout the
[footbet](https://johngavin.github.io/footbet/) package.
See also: [Data Sources](data-sources.html) |
[Data Cleaning](data-cleaning.html) | [EDA](eda.html) |
[Models & Betting](models-betting.html)

## Odds Formats

**Decimal:** Total payout per unit staked (e.g., 2.50 = 1.50 profit + 1.00 stake). Standard in Europe.

**Fractional:** Profit relative to stake (e.g., 3/2 = 3 units profit per 2 staked). UK format.

**American:** Positive = profit on 100 stake (+150 = 150 profit). Negative = stake to win 100 (-200 = bet 200 to win 100).

| Format | Example | Implied Probability |
|--------|---------|---------------------|
| Decimal 2.50 | Win 2.50 per 1 staked | 40% (1/2.50) |
| Fractional 3/2 | Win 3 per 2 staked | 40% (2/(2+3)) |
| American +150 | Win 150 per 100 staked | 40% (100/(100+150)) |
| American -200 | Stake 200 to win 100 | 66.7% (200/(200+100)) |

## Market Types

- **1X2 (Match Result):** Home win (H), Draw (D), Away win (A)
- **Over/Under (O/U):** Total goals vs a line (usually 2.5)
- **Asian Handicap (AH):** Handicap eliminating the draw outcome
- **BTTS:** Both teams to score (yes/no)

## Probability Concepts

- **Implied Probability:** `1 / odds`. For 2.50 odds = 40%.
- **Overround (Vig):** Bookmaker margin. Sum of implied probs minus 100%.
- **Fair Odds:** After removing overround. Methods: multiplicative, power, Shin.
- **Closing Line Value (CLV):** Your odds vs closing odds. See `closing_line_value()`.

## Performance Metrics

- **Log Loss:** $-\frac{1}{n} \sum \log(p_i)$ — penalizes confident wrong predictions
- **Brier Score:** $\frac{1}{n} \sum (p_i - o_i)^2$ — MSE for probabilities
- **RPS:** Ranked probability score for ordered outcomes (H > D > A)
- **ROI:** Total profit / total staked (%)
- **Sharpe Ratio:** Risk-adjusted return. See `betting_sharpe_ratio()`.
- **Maximum Drawdown:** Largest peak-to-trough decline

## Statistical Models

- **Poisson Regression:** Goals ~ Poisson(lambda) with team attack/defense parameters
- **Dixon-Coles:** Poisson + correlation correction for low-scoring outcomes + time decay
- **Expected Goals (xG):** Shot-level goal probability, better predictor than actual goals
- **Elo Rating:** Match-by-match team strength. See `compute_elo()` with COOPER enhancements (dynamic K, margin weighting, league reversion, asymmetric wins)

### Copula Models {#copula}

A [copula](https://en.wikipedia.org/wiki/Copula_(probability_theory)) models the dependence structure between home and away goals separately from their marginal distributions. This is important because:

1. **Independent Poisson underpredicts 0-0 draws.** The observed frequency of 0-0 is ~7%, but independent Poisson(1.5) × Poisson(1.2) predicts ~5%. This excess clustering at (0,0) indicates positive lower-tail dependence.

2. **Home advantage creates asymmetry.** More mass below the diagonal (home > away) than above, violating exchangeability assumed by symmetric copulas.

**Candidate copulas for football scorelines:**

- **[Frank copula](https://en.wikipedia.org/wiki/Copula_(probability_theory)#Most_important_Archimedean_copulas):** Single parameter $\theta$ controlling concordance. Symmetric dependence, captures the general positive association between low-scoring outcomes. $\theta > 0$ means low home goals tend to coincide with low away goals.

- **[Clayton copula](https://en.wikipedia.org/wiki/Clayton_copula):** Parameter $\alpha > 0$ gives strong lower-tail dependence — high probability of both teams scoring zero or one goal simultaneously. This directly models the excess 0-0 and 1-1 scorelines that Dixon-Coles corrects via its $\rho$ parameter.

**Parameterisation for football data:**

$$C(u, v; \theta) = \left(u^{-\theta} + v^{-\theta} - 1\right)^{-1/\theta}$$

where $u = F_H(h)$ and $v = F_A(a)$ are the marginal Poisson CDFs for home and away goals, and $\theta$ is estimated from the observed scoreline frequencies. The Dixon-Coles $\rho$ parameter is effectively a discrete approximation to this copula at the (0,0), (1,0), (0,1), (1,1) cells.

See [issue #43](https://github.com/JohnGavin/footbet/issues/43) for the planned copula investigation and [McHale & Scarf (2011)](https://doi.org/10.1111/j.1467-9876.2011.00782.x) for copula-based football models.

## Betting Strategy

- **Kelly Criterion:** $f^* = \frac{p \cdot b - q}{b}$ — optimal bet sizing
- **Value Bet:** Model probability exceeds market implied probability
- **Edge:** $p \times (o-1) - (1-p)$ — expected profit per unit staked
- **Sharp Bookmaker:** Accurate odds (Pinnacle). **Soft Bookmaker:** Less accurate, slower to adjust.

## Data Terminology {#data-terminology}

- **FTR:** Full Time Result (H/D/A)
- **FTHG / FTAG:** Full Time Home/Away Goals
- **HTHG / HTAG / HTR:** Half Time Home/Away Goals / Result

### Match Statistics {#match-statistics}

- **HS/AS:** Home/Away Shots
- **HST/AST:** Home/Away Shots on Target
- **HC/AC:** Home/Away Corners
- **HF/AF:** Home/Away Fouls
- **HY/AY:** Home/Away Yellow Cards
- **HR/AR:** Home/Away Red Cards

## References

**Package functions:** `?target_leagues`, `?parse_fd_csv`, `?devig_shin`, `?fit_dixon_coles`, `?kelly_stake`, `?closing_line_value`, `?betting_sharpe_ratio`

**External:** [football-data.co.uk](https://www.football-data.co.uk/) | [Pinnacle](https://www.pinnacle.com/) | [Dixon & Coles (1997)](https://doi.org/10.1111/1467-9876.00065) | [goalmodel](https://cran.r-project.org/package=goalmodel)

```{r build-info, echo = FALSE, results = "asis"}
safe_tar_read("vig_build_info")
```
