---
title: "Models & Betting Strategy"
subtitle: "Walk-forward CV, value bets, Kelly staking"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    page-layout: full
    code-fold: true
    include-in-header:
      text: |
        <script src="https://cdn.jsdelivr.net/npm/mermaid/dist/mermaid.min.js"></script>
        <script>
          mermaid.initialize({
            startOnLoad: true,
            securityLevel: 'loose',
            theme: 'dark',
            themeVariables: {
              background: '#000000',
              primaryColor: '#999999',
              primaryTextColor: '#000000',
              primaryBorderColor: '#CC0000',
              lineColor: '#CC0000',
              secondaryColor: '#333333',
              tertiaryColor: '#333333'
            }
          });
        </script>
vignette: >
  %\VignetteIndexEntry{Models & Betting Strategy}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(htmltools)
library(DT)
library(dplyr)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) {
    rds_path <- system.file(file.path("extdata", "vignettes", paste0(name, ".rds")), package = "footbet")
    if (!nzchar(rds_path) || !file.exists(rds_path))
      rds_path <- here::here("inst", "extdata", "vignettes", paste0(name, ".rds"))
    if (file.exists(rds_path)) result <- readRDS(rds_path)
  }
  if (is.null(result)) {
    htmltools::div(style = "background: #dc3545; color: white; padding: 1em; border-radius: 4px;",
      htmltools::p(paste0("[MISSING EVIDENCE] Target `", name, "` not found.")))
  } else result
}
```

```{=html}
<link rel="stylesheet" href="https://cdn.datatables.net/1.13.8/css/jquery.dataTables.min.css">
<script src="https://cdn.datatables.net/1.13.8/js/jquery.dataTables.min.js"></script>
<style>
table.dataTable tbody td, table.dataTable tbody th,
.table tbody td, .table tbody th,
table td, table th { color: #ffffff !important; }
table.dataTable thead th, table.dataTable thead td,
.table thead th, .table thead td,
table thead th { color: #ffffff !important; font-weight: bold !important; }
.dataTables_info, .dataTables_paginate, .dataTables_length,
.dataTables_filter label, .dataTables_wrapper { color: #ffffff !important; }
.dataTables_filter input, .dataTables_length select {
  color: #ffffff !important; background-color: #333 !important; }
caption, .dataTables_wrapper caption { color: #cccccc !important; }
table.dataTable, .table, table { background-color: #1a1a1a !important; }
table.dataTable tbody tr, .table tbody tr { background-color: #1a1a1a !important; }
table.dataTable tbody tr:hover { background-color: #333 !important; }
</style>
```

Part 4 of the [footbet](https://johngavin.github.io/footbet/) analytics guide.
See also: [Data Sources](data-sources.html) |
[Data Cleaning](data-cleaning.html) | [EDA](eda.html) |
[Glossary](glossary.html)

::: {.panel-tabset}

## Models

### Cross-Validation

Football data is time-ordered, so standard k-fold CV would leak future
information. The pipeline uses **walk-forward** (expanding window) splits:

- **Training window**: 24 months of historical matches
- **Test window**: 1 month of subsequent matches
- The window slides forward by 1 month for each fold

```{r vig-cv-diagram}
#| echo: false
#| output: asis
safe_tar_read("vig_cv_html")
```

*Walk-forward cross-validation diagram. X-axis: time (months). Y-axis: CV fold index. Blue shading = training window (24 months), orange shading = test window (1 month). Each row represents one fold, with the window sliding forward by one month. This design prevents future data from leaking into model training --- a critical requirement for time-series prediction. Source: [footbet](https://johngavin.github.io/footbet/) pipeline. See [Glossary > Cross-Validation](glossary.html).*

### GLM Poisson

The baseline model is a
[Poisson regression](https://en.wikipedia.org/wiki/Poisson_regression)
for goal counts, following Maher (1982):

$$\text{goals}_{ij} \sim \text{Poisson}(\lambda_{ij})$$

$$\log(\lambda_{ij}) = \mu + \alpha_i + \beta_j + \gamma \cdot \mathbb{1}[\text{home}]$$

where $\alpha_i$ is team $i$'s attack strength, $\beta_j$ is team $j$'s
defence weakness, and $\gamma$ captures home advantage.

### Dixon-Coles

Dixon & Coles (1997) extend the Poisson model with:

1. **Correlation correction**: A parameter $\rho$ adjusts the joint
   probability of low-scoring outcomes (0-0, 1-0, 0-1, 1-1)
2. **Time-decay weights**: Recent matches receive higher weight via
   exponential decay ($\xi = 0.003$, ~1-year half-life)

### CV Metrics

Log-loss, Brier score, and RPS per CV fold for both models, with Pinnacle
as benchmark. All are **proper scoring rules** --- lower is better.

```{r vig-cv-metrics-plot}
safe_tar_read("vig_cv_metrics_plot")
```

*Cross-validation scoring metrics by fold. X-axis: CV fold number. Y-axis: metric value (lower = better). Lines: GLM Poisson (blue), Dixon-Coles (orange), Pinnacle benchmark (grey dashed). Three panels show log-loss, Brier score, and RPS respectively. Dixon-Coles slightly outperforms GLM on most folds, but both trail Pinnacle closing odds. Pinnacle's sharp market remains the hardest benchmark to beat. Source: walk-forward CV on [football-data.co.uk](https://www.football-data.co.uk/) match data. See [Glossary > Performance Metrics](glossary.html).*

```{r vig-model-comparison-test}
safe_tar_read("vig_model_comparison_test")
```

*Statistical comparison of model performance. Columns: model name, mean metric, confidence interval, p-value. This table tests whether Dixon-Coles significantly outperforms GLM Poisson on log-loss. While Dixon-Coles shows a small improvement, the difference is often not statistically significant. Neither model consistently beats the Pinnacle closing line. Source: paired t-test across CV folds using [football-data.co.uk](https://www.football-data.co.uk/) data. See [Glossary > Log Loss](glossary.html).*

::: {.callout-note}
**Key Finding:** Dixon-Coles slightly outperforms GLM. Both models trail
Pinnacle on most folds.
:::

### Model Comparison

```{r vig-model-comparison-table}
model_comp <- safe_tar_read("vig_model_comparison_table")
if (is.data.frame(model_comp)) {
  model_comp |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Model comparison. Lower scores = better. Positive edge = outperforms Pinnacle."
      ),
      options = list(pageLength = 10, dom = "t")
    )
} else {
  model_comp
}
```

*Summary model comparison table. Columns: model (GLM/Dixon-Coles/Pinnacle), mean log-loss, mean Brier score, mean RPS, edge vs Pinnacle (%). Lower scoring metrics indicate better calibration. Positive edge means the model outperforms Pinnacle on average. Dixon-Coles achieves the lowest log-loss among the fitted models. The edge column quantifies how close each model comes to the sharp-market benchmark. Source: aggregated CV results from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Performance Metrics](glossary.html).*

## Betting Strategy

### Value Bets

A **value bet** exists when the model's estimated probability exceeds the
market's implied probability by a minimum edge threshold:

- **Minimum edge**: 3% (model_prob - market_prob > 0.03)
- **Odds filter**: Only bets between 1.50 and 10.00 decimal odds
- **Devigged odds**: Market probabilities via Shin (1993) method

```{r vig-value-bets-summary}
vb_summary <- safe_tar_read("vig_value_bets_summary")
if (is.data.frame(vb_summary)) {
  vb_summary |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Value bet summary by outcome and league. Edge = model prob - market prob."
      ),
      filter = "top",
      options = list(pageLength = 15, dom = "Bfrtip")
    )
} else {
  vb_summary
}
```

*Value bet summary table. Columns: league code, outcome (H/D/A), number of value bets identified, mean edge (%), hit rate (%), ROI (%). Edge = model probability minus Shin-devigged market probability. Home value bets tend to be more frequent; draw value bets show the highest mean edge but lowest hit rate. Away value bets are rarest due to lower model confidence. Source: Dixon-Coles model applied to [football-data.co.uk](https://www.football-data.co.uk/) with Pinnacle odds. See [Glossary > Value Bet](glossary.html) and [Glossary > Edge](glossary.html).*

### Edge Distribution

```{r vig-edge-distribution}
safe_tar_read("vig_edge_distribution")
```

*Distribution of betting edges across all identified value bets. X-axis: edge size (%, model probability minus market probability). Y-axis: frequency count. The histogram shows most edges cluster between 3-8%, with a long right tail. Edges above 15% are rare and may indicate model mis-specification rather than genuine value. The 3% minimum threshold filters out noise. Source: Dixon-Coles predictions vs Pinnacle devigged odds from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Edge](glossary.html) and [Glossary > Fair Odds](glossary.html).*

::: {.callout-note}
**Key Finding:** Most edges cluster at 3-8%. Large edges (>15%) are rare and
may signal model error.
:::

### Kelly Staking

The pipeline uses **quarter Kelly** staking with guardrails:

$$f^* = \frac{1}{4} \cdot \frac{p \cdot (b + 1) - 1}{b}$$

- **Max stake**: 3% of current bankroll per bet
- **Drawdown halt**: Stakes halved when bankroll drops 20% from peak

```{r vig-kelly-diagram}
#| echo: false
#| output: asis
safe_tar_read("vig_kelly_html")
```

*Kelly criterion staking flowchart. The diagram shows the decision tree: compute full Kelly fraction, apply quarter-Kelly scaling, cap at 3% of bankroll, then check the drawdown guard (if bankroll is 20%+ below peak, halve the stake). This risk management pipeline prevents catastrophic losses from model error while still exploiting value bets. Source: [footbet](https://johngavin.github.io/footbet/) staking module. See [Glossary > Kelly Criterion](glossary.html).*

```{r vig-kelly-stake-distribution}
safe_tar_read("vig_kelly_stake_distribution")
```

*Distribution of Kelly stake sizes. X-axis: stake as percentage of bankroll. Y-axis: frequency count. Most stakes fall between 0.5-2% of bankroll due to the quarter-Kelly scaling and 3% cap. The distribution is right-skewed, with very few bets reaching the 3% maximum. Smaller stakes correspond to lower-edge bets. Source: simulated staking on [football-data.co.uk](https://www.football-data.co.uk/) value bets. See [Glossary > Kelly Criterion](glossary.html) and [Glossary > Edge](glossary.html).*

## P&L

### P&L Simulation

The bankroll curve below shows the simulated profit & loss trajectory for all
GLM value bets, staked with quarter Kelly and drawdown guardrails.

```{r vig-pnl-curve}
safe_tar_read("vig_pnl_curve")
```

*Cumulative profit & loss trajectory over time. X-axis: date (match timeline). Y-axis: bankroll value (units, starting at 100). The line tracks the running bankroll after each settled bet. Upward slopes indicate profitable periods; downward slopes are drawdown periods. The trajectory depends on model accuracy and the variance inherent in sports betting. Key risk metric is the maximum peak-to-trough decline. Source: quarter-Kelly simulation on [football-data.co.uk](https://www.football-data.co.uk/) value bets. See [Glossary > ROI](glossary.html) and [Glossary > Maximum Drawdown](glossary.html).*

### Drawdown

```{r vig-drawdown-plot}
safe_tar_read("vig_drawdown_plot")
```

*Drawdown plot showing peak-to-trough declines over time. X-axis: date (match timeline). Y-axis: drawdown percentage (0% = at peak, negative = below peak). Red shading indicates drawdown depth. The deepest trough represents the maximum drawdown --- the worst-case loss from peak bankroll. Extended drawdown periods test bankroll management discipline. Source: quarter-Kelly P&L simulation using [football-data.co.uk](https://www.football-data.co.uk/) data. See [Glossary > Maximum Drawdown](glossary.html).*

::: {.callout-note}
**Key Finding:** Trajectory depends on model accuracy and market conditions.
Maximum drawdown is the key risk metric.
:::

### Summary

```{r vig-pnl-summary-table}
pnl_summary <- safe_tar_read("vig_pnl_summary_table")
if (is.data.frame(pnl_summary)) {
  pnl_summary |>
    DT::datatable(
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "P&L summary. ROI = total profit / total staked. Sharpe = risk-adjusted return."
      ),
      options = list(pageLength = 10, dom = "t")
    )
} else {
  pnl_summary
}
```

*P&L summary statistics table. Columns: total bets, total staked (units), total profit (units), ROI (%), Sharpe ratio, maximum drawdown (%), win rate (%). ROI measures overall profitability; the Sharpe ratio adjusts returns for volatility. Maximum drawdown quantifies worst-case risk exposure. A Sharpe above 1.0 suggests strong risk-adjusted performance. Source: end-to-end simulation on [football-data.co.uk](https://www.football-data.co.uk/) value bets with quarter-Kelly staking. See [Glossary > ROI](glossary.html), [Glossary > Sharpe Ratio](glossary.html), and [Glossary > Maximum Drawdown](glossary.html).*

:::
