---
title: "Model Fitting & Evaluation"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: false
    page-layout: full
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Model Fitting & Evaluation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE
)
library(targets)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) invisible(NULL) else result
}
```

This vignette covers the modelling pipeline: walk-forward cross-validation,
two statistical models, and benchmarking against Pinnacle implied probabilities.

::: {.panel-tabset}

## Cross-Validation

Football data is time-ordered, so standard k-fold CV would leak future
information. The pipeline uses **walk-forward** (expanding window) splits:

- **Training window**: 24 months of historical matches
- **Test window**: 1 month of subsequent matches
- The window slides forward by 1 month for each fold

This mimics a realistic betting scenario where you train on past data and
predict the next month.

`footbet::walk_forward_splits()` generates the date-based train/test splits,
and `evaluate_glm_baseline()` / `evaluate_dc()` iterate over them.

## GLM Poisson

The baseline model is a
[Poisson regression](https://en.wikipedia.org/wiki/Poisson_regression)
for goal counts, following Maher (1982):

$$\text{goals}_{ij} \sim \text{Poisson}(\lambda_{ij})$$

$$\log(\lambda_{ij}) = \mu + \alpha_i + \beta_j + \gamma \cdot \mathbb{1}[\text{home}]$$

where $\alpha_i$ is team $i$'s attack strength, $\beta_j$ is team $j$'s
defence weakness, and $\gamma$ captures home advantage. The model assumes
independence between home and away goals.

`footbet::fit_poisson_glm()` fits this model; `predict_glm()` generates a
bivariate score probability matrix (up to 10x10 goals) and derives 1X2,
over/under, and Asian handicap probabilities.

## Dixon-Coles

Dixon & Coles (1997) extend the Poisson model with:

1. **Correlation correction**: A parameter $\rho$ adjusts the joint
   probability of low-scoring outcomes (0-0, 1-0, 0-1, 1-1) where the
   independence assumption fails most
2. **Time-decay weights**: Recent matches receive higher weight via an
   exponential decay parameter $\xi$

The pipeline uses $\xi = 0.003$ (approximately 1-year half-life) via the
[goalmodel](https://cran.r-project.org/package=goalmodel) package.

## CV Metrics

The plot below shows
[log-loss](https://en.wikipedia.org/wiki/Cross-entropy),
[Brier score](https://en.wikipedia.org/wiki/Brier_score), and
[RPS](https://en.wikipedia.org/wiki/Ranked_probability_score)
per CV fold for both models, with Pinnacle as a benchmark.
All three are **proper scoring rules** — lower is better.

```{r vig-cv-metrics-plot}
safe_tar_read("vig_cv_metrics_plot")
```

**Key findings:**

- Dixon-Coles slightly outperforms GLM
- Both models trail Pinnacle on most folds
- Pinnacle's edge is consistent across metrics

## Model Comparison

The table below summarises mean metrics across all folds, with the Pinnacle
benchmark for comparison. Positive "edge" means the model outperforms Pinnacle.

```{r vig-model-comparison-table}
safe_tar_read("vig_model_comparison_table")
```

## Value Bets

A **value bet** exists when the model's estimated probability exceeds the
market's implied probability by a minimum edge threshold. The pipeline uses:

- **Minimum edge**: 3% (model_prob - market_prob > 0.03)
- **Odds filter**: Only bets between 1.50 and 10.00 decimal odds
- **Devigged odds**: Market probabilities via [Shin (1993)](https://doi.org/10.2307/2234526) method

```{r vig-value-bets-summary}
safe_tar_read("vig_value_bets_summary")
```

```{r vig-edge-distribution}
safe_tar_read("vig_edge_distribution")
```

**Key findings:**

- Most edges cluster at 3-8%
- Large edges (>15%) are rare and may signal model error

## Kelly Staking

The pipeline uses **quarter Kelly** staking:

$$f^* = \frac{1}{4} \cdot \frac{p \cdot (b + 1) - 1}{b}$$

where $p$ is the model probability and $b$ is the decimal odds minus 1.

**Guardrails:**

- **Max stake**: 3% of current bankroll per bet
- **Drawdown halt**: Stakes halved when bankroll drops 20% from peak
- **`apply_guardrails()`**: Automatically reduces stake size during drawdowns

```{r vig-kelly-stake-distribution}
safe_tar_read("vig_kelly_stake_distribution")
```

## P&L Simulation

The bankroll curve below shows the simulated profit & loss trajectory for all
GLM value bets, staked with quarter Kelly and drawdown guardrails.
Use the range slider to zoom into specific periods.

```{r vig-pnl-curve}
safe_tar_read("vig_pnl_curve")
```

```{r vig-drawdown-plot}
safe_tar_read("vig_drawdown_plot")
```

**Key findings:**

- Trajectory depends on model accuracy and market conditions
- Drawdown guardrails limit losses during poor runs
- Maximum drawdown is the key risk metric

### Summary

```{r vig-pnl-summary-table}
safe_tar_read("vig_pnl_summary_table")
```

:::
