---
title: "Models & Betting Strategy"
subtitle: "Walk-forward CV, value bets, Kelly staking"
format:
  html:
    theme: darkly
    toc: false
    page-layout: full
    code-fold: true
    include-in-header:
      text: |
        <script>document.documentElement.setAttribute('data-bs-theme','dark');</script>
        <style>html,body{background:#000!important;color:#fff!important;}</style>
        <script src="https://cdn.jsdelivr.net/npm/mermaid/dist/mermaid.min.js"></script>
        <script>
          mermaid.initialize({
            startOnLoad: true,
            securityLevel: 'loose',
            theme: 'dark',
            themeVariables: {
              background: '#000000',
              primaryColor: '#999999',
              primaryTextColor: '#000000',
              primaryBorderColor: '#CC0000',
              lineColor: '#CC0000',
              secondaryColor: '#333333',
              tertiaryColor: '#333333'
            }
          });
        </script>
vignette: >
  %\VignetteIndexEntry{Models & Betting Strategy}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(htmltools)
library(DT)
library(dplyr)

if (file.exists(here::here("_targets"))) {
  tar_config_set(store = here::here("_targets"))
}

safe_tar_read <- function(name) {
  result <- tryCatch(targets::tar_read_raw(name), error = function(e) NULL)
  if (is.null(result)) {
    rds_path <- system.file(file.path("extdata", "vignettes", paste0(name, ".rds")), package = "footbet")
    if (!nzchar(rds_path) || !file.exists(rds_path))
      rds_path <- here::here("inst", "extdata", "vignettes", paste0(name, ".rds"))
    if (file.exists(rds_path)) result <- readRDS(rds_path)
  }
  if (is.null(result)) {
    htmltools::div(style = "background: #dc3545; color: white; padding: 1em; border-radius: 4px;",
      htmltools::p(paste0("[MISSING EVIDENCE] Target `", name, "` not found.")))
  } else result
}

dark_dt <- function(data, caption = "", ...) {
  nums <- sapply(data, is.numeric)
  data[nums] <- lapply(data[nums], function(x) {
    if (all(x == floor(x), na.rm = TRUE)) x  # integers stay as-is
    else if (max(abs(x), na.rm = TRUE) > 1) round(x, 3)  # scores
    else round(x, 4)  # probabilities
  })
  DT::datatable(
    data,
    caption = caption,
    options = list(
      pageLength = 20,
      dom = "t",
      scrollX = TRUE,
      initComplete = DT::JS(
        "function(settings, json) {",
        "  $(this.api().table().container()).css({'background-color': '#1a1a2e', 'color': '#e0e0e0'});",
        "  $(this.api().table().header()).css({'background-color': '#16213e', 'color': '#e0e0e0'});",
        "  $('td', this.api().table().body()).css({'background-color': '#1a1a2e', 'color': '#e0e0e0'});",
        "}"
      )
    ),
    escape = FALSE,
    rownames = FALSE,
    ...
  )
}
```

```{=html}
<link rel="stylesheet" href="https://cdn.datatables.net/1.13.8/css/jquery.dataTables.min.css">
<script src="https://cdn.datatables.net/1.13.8/js/jquery.dataTables.min.js"></script>
<script>document.documentElement.setAttribute('data-bs-theme', 'dark');</script>
<style>
body, .page-body, main, article, .contents { background-color: #000000 !important; color: #ffffff !important; }
table, .table, table.dataTable { background-color: #000000 !important; color: #ffffff !important; }
table td, table th, .table td, .table th { color: #ffffff !important; background-color: #000000 !important; border-color: #333 !important; }
table thead th, .table thead th { color: #ffffff !important; font-weight: bold !important; background-color: #111 !important; border-color: #333 !important; }
table tbody tr:hover td { background-color: #222 !important; }
.dataTables_info, .dataTables_paginate, .dataTables_length,
.dataTables_filter label, .dataTables_wrapper { color: #ffffff !important; }
.dataTables_filter input, .dataTables_length select { color: #ffffff !important; background-color: #333 !important; border-color: #555 !important; }
caption, .dataTables_wrapper caption { color: #cccccc !important; }
.dataTables_wrapper, .cell-output-display { overflow-x: auto !important; }
table { white-space: nowrap; }
p, li, h1, h2, h3, h4, h5, h6, span, em, strong, a, code { color: #ffffff !important; }
a:hover { color: #66b3ff !important; }
code { background-color: #222 !important; }
.callout { background-color: #111 !important; border-color: #333 !important; }
</style>
```

Part 4 of the [footbet](https://johngavin.github.io/footbet/) analytics guide.
See also: [Data Sources](data-sources.html) |
[Data Cleaning](data-cleaning.html) | [EDA](eda.html) |
[Glossary](glossary.html)

---

## Leaderboard

All models evaluated in footbet, ranked by market and performance.
1X2 models are scored on calibration (log-loss, Brier, RPS).
AH models are scored on walk-forward P&L (ROI%, Sharpe, max drawdown).
See also the [AH Backtest vignette](ah-backtest.html) for equity curves and per-season detail.

::: {.callout-note}
Brier scores for 3-outcome football (1X2) range from 0 (perfect) to 2 (worst). A naive model predicting equal 1/3 probabilities scores 0.667. All models here score below 0.667, confirming they outperform random guessing — though the margin is slim.
:::

:::{.panel-tabset}

### Calibration (1X2)

```{r vig-leaderboard-calibration, echo = FALSE}
lb_cal <- safe_tar_read("vig_leaderboard_calibration")
if (is.data.frame(lb_cal) && nrow(lb_cal) > 0) {
  lb_cal_dt <- lb_cal |>
    dplyr::mutate(
      model = dplyr::case_when(
        model == "GLM Poisson"     ~ "<a href='#glm-poisson'>GLM Poisson</a>",
        model == "Dixon-Coles"     ~ "<a href='#dixon-coles'>Dixon-Coles</a>",
        model == "Pinnacle closing" ~ "Pinnacle closing",
        TRUE ~ model
      )
    )
  dark_dt(
    lb_cal_dt,
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left;",
      paste0(
        "1X2 calibration leaderboard (", nrow(lb_cal), " models). ",
        "Metrics from walk-forward CV (lower = better). ",
        "Pinnacle closing odds are the benchmark all models must beat. ",
        "Model names link to their detail sections. ",
        "Source: plan_evaluation.R, plan_brms_1x2.R."
      )
    )
  )
} else {
  lb_cal
}
```

*1X2 model calibration. Columns: model (links to detail section), log-loss, Brier score, RPS (all lower = better), number of CV folds. brms hierarchical achieves the best log-loss among fitted models, but all trail Pinnacle closing odds. Source: walk-forward CV on [football-data.co.uk](https://www.football-data.co.uk/). See [MODEL_CATALOGUE.md](https://github.com/JohnGavin/footbet/blob/main/plans/MODEL_CATALOGUE.md).*

### P&L (AH)

```{r vig-leaderboard-pnl, echo = FALSE}
lb_pnl <- safe_tar_read("vig_leaderboard_pnl")
if (is.data.frame(lb_pnl) && nrow(lb_pnl) > 0) {
  dark_dt(
    lb_pnl,
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left;",
      paste0(
        "AH P&L leaderboard (", nrow(lb_pnl), " models, flat staking). ",
        "Walk-forward backtest over 6 seasons (19-20 to 24-25). ",
        "No model achieves persistent positive ROI. ",
        "Source: plan_ranger_1x2.R, plan_vignette_outputs.R."
      )
    )
  )
} else {
  lb_pnl
}
```

*AH walk-forward P&L leaderboard. Columns: model, ROI% (flat staking), number of bets, Sharpe ratio, maximum drawdown, win rate. GLM-Ranger intersection is least bad at +0.3% ROI over 2,106 bets. No systematic edge exists against Pinnacle. Source: [football-data.co.uk](https://www.football-data.co.uk/) AH odds. See [AH Backtest vignette](ah-backtest.html) for equity curves and per-season breakdowns.*

:::

---

## Cross-Validation

Football data is time-ordered, so standard k-fold CV would leak future
information. The pipeline uses **walk-forward** (expanding window) splits:

- **Training window**: 24 months of historical matches
- **Test window**: 1 month of subsequent matches
- The window slides forward by 1 month for each fold

:::{.panel-tabset}

### Walk-Forward Design

```{r vig-cv-diagram}
#| echo: false
#| output: asis
safe_tar_read("vig_cv_html")
```

*Walk-forward cross-validation diagram. X-axis: time (months). Y-axis: CV fold index. Blue shading = training window (24 months), orange shading = test window (1 month). Each row represents one fold, with the window sliding forward by one month. This design prevents future data from leaking into model training --- a critical requirement for time-series prediction. Source: [footbet](https://johngavin.github.io/footbet/) pipeline. See [Glossary > Cross-Validation](glossary.html).*

#### GLM Poisson {#glm-poisson}

The baseline model is a
[Poisson regression](https://en.wikipedia.org/wiki/Poisson_regression)
for goal counts, following Maher (1982):

$$\text{goals}_{ij} \sim \text{Poisson}(\lambda_{ij})$$

$$\log(\lambda_{ij}) = \mu + \alpha_i + \beta_j + \gamma \cdot \mathbb{1}[\text{home}]$$

where $\alpha_i$ is team $i$'s attack strength, $\beta_j$ is team $j$'s
defence weakness, and $\gamma$ captures home advantage.

#### Dixon-Coles {#dixon-coles}

Dixon & Coles (1997) extend the Poisson model with:

1. **Correlation correction**: A parameter $\rho$ adjusts the joint
   probability of low-scoring outcomes (0-0, 1-0, 0-1, 1-1)
2. **Time-decay weights**: Recent matches receive higher weight via
   exponential decay ($\xi = 0.003$, ~1-year half-life)

### Scoring Metrics

Log-loss, Brier score, and RPS per CV fold for both models, with Pinnacle
as benchmark. All are **proper scoring rules** --- lower is better.

```{r vig-cv-metrics-plot}
safe_tar_read("vig_cv_metrics_plot")
```

*Cross-validation scoring metrics by fold. Three panels (log-loss, Brier, RPS), each showing one metric across all walk-forward folds. X-axis: CV fold number. Y-axis: metric value (lower = better). Dots: GLM Poisson (blue), Dixon-Coles (purple), Pinnacle benchmark (grey dashed). Dixon-Coles slightly outperforms GLM on most folds, but both trail Pinnacle closing odds. Legend entries are clickable — click to show/hide individual models. Source: walk-forward CV on [football-data.co.uk](https://www.football-data.co.uk/) match data. See [Glossary > Performance Metrics](glossary.html).*

```{r vig-model-comparison-test}
safe_tar_read("vig_model_comparison_test")
```

*Statistical comparison of model performance. Columns: model name, mean metric, confidence interval, p-value. This table tests whether Dixon-Coles significantly outperforms GLM Poisson on log-loss. While Dixon-Coles shows a small improvement, the difference is often not statistically significant. Neither model consistently beats the Pinnacle closing line. Source: paired t-test across CV folds using [football-data.co.uk](https://www.football-data.co.uk/) data. See [Glossary > Log Loss](glossary.html).*

::: {.callout-note}
**Key Finding:** Dixon-Coles slightly outperforms GLM. Both models trail
Pinnacle on most folds.
:::

### Model Comparison

```{r vig-model-comparison-table}
model_comp <- safe_tar_read("vig_model_comparison_table")
if (is.data.frame(model_comp)) {
  dark_dt(
    model_comp,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Model comparison (3 d.p.). log_loss, brier, rps: lower = better; all are proper scoring rules but are not directly comparable across models fitted on different data subsets. edge: model mean log-loss minus Pinnacle mean log-loss (positive = worse than Pinnacle)."
    )
  )
} else {
  model_comp
}
```

*Summary model comparison table. Columns: model (GLM/Dixon-Coles/Pinnacle), mean log-loss, mean Brier score, mean RPS, edge vs Pinnacle (%). Lower scoring metrics indicate better calibration. Positive edge means the model outperforms Pinnacle on average. Dixon-Coles achieves the lowest log-loss among the fitted models. The edge column quantifies how close each model comes to the sharp-market benchmark. Source: aggregated CV results from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Performance Metrics](glossary.html).*

:::

---

## Betting Strategy

:::{.panel-tabset}

### Value Bets

A **value bet** exists when the model's estimated probability exceeds the
market's implied probability by a minimum edge threshold:

- **Minimum edge**: 3% (model_prob - market_prob > 0.03)
- **Odds filter**: Only bets between 1.50 and 10.00 decimal odds
- **Devigged odds**: Market probabilities via Shin (1993) method

```{r vig-value-bets-summary}
vb_summary <- safe_tar_read("vig_value_bets_summary")
if (is.data.frame(vb_summary)) {
  dark_dt(
    vb_summary,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Value bet summary. Columns: outcome (H/D/A = Home/Draw/Away win), league_code, n_bets (count of qualifying bets), mean_edge (mean of model_prob - devigged_market_prob), mean_odds (mean decimal odds at bet time), mean_kelly (mean quarter-Kelly stake fraction). Edge threshold: >3%. Odds filter: 1.50-10.00."
    )
  )
} else {
  vb_summary
}
```

*Value bet summary table. Columns: league code, outcome (H/D/A), number of value bets identified, mean edge (%), hit rate (%), ROI (%). Edge = model probability minus Shin-devigged market probability. Home value bets tend to be more frequent; draw value bets show the highest mean edge but lowest hit rate. Away value bets are rarest due to lower model confidence. Source: Dixon-Coles model applied to [football-data.co.uk](https://www.football-data.co.uk/) with Pinnacle odds. See [Glossary > Value Bet](glossary.html) and [Glossary > Edge](glossary.html).*

### Edge Distribution

```{r vig-edge-distribution}
safe_tar_read("vig_edge_distribution")
```

*Distribution of betting edges across all identified value bets. X-axis: edge size (model probability minus market probability). Y-axis: frequency count. The histogram shows most edges cluster between 3-8%, with a long right tail. Edges above 15% are rare and may indicate model mis-specification rather than genuine value. The 3% minimum threshold filters out noise. Legend entries are clickable — click to show/hide individual outcomes. Source: Dixon-Coles predictions vs Pinnacle devigged odds from [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Edge](glossary.html) and [Glossary > Fair Odds](glossary.html).*

```{r vig-edge-panels}
safe_tar_read("vig_edge_panels")
```

*Edge distribution split by outcome (Home/Draw/Away). Three panels share the x-axis (edge size). Summary statistics (mean, median, count above 15%) appear in each panel. Draw edges tend to be higher as the market is less efficient for draws. Source: Dixon-Coles predictions vs Pinnacle devigged odds. See [Glossary > Edge](glossary.html).*

::: {.callout-note}
**Key Finding:** Most edges cluster at 3-8%. Large edges (>15%) are rare and
may signal model error.
:::

### Kelly Staking

The pipeline uses **quarter Kelly** staking with guardrails:

$$f^* = \frac{1}{4} \cdot \frac{p \cdot (b + 1) - 1}{b}$$

- **Max stake**: 3% of current bankroll per bet
- **Drawdown halt**: Stakes halved when bankroll drops 20% from peak

```{r vig-kelly-diagram}
#| echo: false
#| output: asis
safe_tar_read("vig_kelly_html")
```

*Kelly criterion staking flowchart. The diagram shows the decision tree: compute full Kelly fraction, apply quarter-Kelly scaling, cap at 3% of bankroll, then check the drawdown guard (if bankroll is 20%+ below peak, halve the stake). This risk management pipeline prevents catastrophic losses from model error while still exploiting value bets. Source: [footbet](https://johngavin.github.io/footbet/) staking module. See [Glossary > Kelly Criterion](glossary.html).*

```{r vig-kelly-stake-distribution}
safe_tar_read("vig_kelly_stake_distribution")
```

*Distribution of Kelly stake sizes. X-axis: stake as percentage of bankroll. Y-axis: frequency count. Most stakes fall between 0.5-2% of bankroll due to the quarter-Kelly scaling and 3% cap. The distribution is right-skewed, with very few bets reaching the 3% maximum. Smaller stakes correspond to lower-edge bets. Source: simulated staking on [football-data.co.uk](https://www.football-data.co.uk/) value bets. See [Glossary > Kelly Criterion](glossary.html) and [Glossary > Edge](glossary.html).*

::: {.callout-note}
## Cap Impact
Approximately one-third of value bets hit the 3% cap, meaning the Kelly criterion provides minimal practical variation in bet sizing for those bets. See [issue #93](https://github.com/JohnGavin/footbet/issues/93) for alternatives including tiered caps and fractional Kelly approaches.
:::

:::

---

## P&L & Risk

:::{.panel-tabset}

### Equity Curve

The bankroll curve below shows the simulated profit & loss trajectory for all
GLM value bets under realistic assumptions: 2% transaction cost, 1% odds
slippage, flat £10 stakes.

```{r vig-pnl-curve-comparison}
safe_tar_read("vig_pnl_curve_realistic")
```

*GLM Poisson flat-stake P&L. Simulated bankroll (realistic scenario). Y-axis: log scale. Assumptions: 2% transaction cost, 1% odds slippage, flat £10 per bet. Starting bankroll: 1,000. Early volatility reflects smaller sample and higher variance in initial seasons. Recent flatness suggests the market has become more efficient, with Pinnacle odds leaving less exploitable edge. Source: GLM value bets on [football-data.co.uk](https://www.football-data.co.uk/). See [Glossary > Kelly Criterion](glossary.html) and [issue #66](https://github.com/JohnGavin/footbet/issues/66).*

### Drawdown

```{r vig-drawdown-plot}
safe_tar_read("vig_drawdown_plot")
```

*GLM Poisson strategy drawdown over time. X-axis: date. Y-axis: drawdown percentage (0% = at peak, negative = below peak). Red shading = drawdown depth. The deepest trough represents the maximum drawdown — worst-case loss from peak bankroll. Source: quarter-Kelly P&L simulation on [football-data.co.uk](https://www.football-data.co.uk/) data. See [Glossary > Maximum Drawdown](glossary.html).*

::: {.callout-warning}
**Caveat:** Real-world constraints not fully captured here include account limits
on profitable bettors and backtest overfitting. The 2% cost + 1% slippage
approximates Pinnacle's spread. See [issue #66](https://github.com/JohnGavin/footbet/issues/66).
:::

### Summary

```{r vig-pnl-summary-table}
pnl_summary <- safe_tar_read("vig_pnl_summary_table")
if (is.data.frame(pnl_summary)) {
  dark_dt(
    pnl_summary,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "P&L summary: realistic scenario (2% cost, 1% slippage, flat £10 stakes). Rows: one metric per row."
    )
  )
} else {
  pnl_summary
}
```

*P&L summary (realistic scenario). See [issue #66](https://github.com/JohnGavin/footbet/issues/66) for assumptions.*

:::

---

## Out-of-Sample

:::{.panel-tabset}

All results in this vignette are out-of-sample. Two evaluation frameworks
are in use and it is worth distinguishing them:

**Framework A — fixed split** (`plan_oos.R`): Train on seasons 2015–2020,
validate on 2021–2023. The P&L simulation above uses this split. The test
period (2024–2026) is reserved for a final one-shot evaluation
([issue #67](https://github.com/JohnGavin/footbet/issues/67)).

**Framework C — rolling window** (`plan_ranger_1x2.R`, `plan_xgboost_1x2.R`):
24-month rolling training window, 1-month test, slides forward by one month
for each fold. The CV metrics table above uses Framework C. Both frameworks
are out-of-sample; they differ in how training data accumulates over time.

The leaderboard combines both: calibration metrics (log-loss, Brier, RPS)
come from Framework C walk-forward CV; P&L metrics (ROI, Sharpe, max DD)
come from Framework A fixed-split backtests.

### OOS Comparison

```{r vig-oos-comparison}
oos <- safe_tar_read("vig_oos_comparison")
if (is.data.frame(oos)) {
  dark_dt(
    oos,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Framework A OOS P&L: train 2015-2020, validate 2021-2023. Realistic = 2% cost, 1% slippage, flat £10 stakes."
    )
  )
} else {
  oos
}
```

*Out-of-sample evaluation (Framework A: fixed split). Train 2015–2020, validate 2021–2023. 2% transaction cost, 1% slippage, flat £10 stakes. The edge from the model (~5-10%) is consumed by the bookmaker's margin (~2-4%) plus slippage.*

### OOS by League

```{r vig-oos-by-league}
oos_league <- safe_tar_read("vig_oos_by_league")
if (is.data.frame(oos_league)) {
  dark_dt(
    oos_league,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "OOS ROI by league. Flat £10, 2% cost, 1% slippage. No league is profitable."
    )
  )
} else {
  oos_league
}
```

*Out-of-sample ROI by league. E0 (Premier League) is closest to breakeven at -2.1%, while I2 (Serie B) is worst at -20.1%. No league is profitable after costs. Top 5 and 2nd Tier leagues show similar patterns. The variation in ROI across leagues suggests league-specific models or features might improve performance, but the fundamental barrier is the Pinnacle spread. See [issue #67](https://github.com/JohnGavin/footbet/issues/67) for the test-period evaluation and [issue #43](https://github.com/JohnGavin/footbet/issues/43) for copula model investigation.*

### xG Calibration

Rolling xG features improve 1X2 calibration by ~7% log-loss over a goals-only GLM baseline in walk-forward CV. This improvement survives the 7-day bet-time cutoff leakage correction unchanged (cut7 vs cut0 Brier differs by 0.001). The effect is real, not a leakage artefact.

```{r vig-xg-per-league, echo = FALSE}
xg_pl <- safe_tar_read("vig_xg_per_league")
if (is.data.frame(xg_pl) && nrow(xg_pl) > 0) {
  dark_dt(
    xg_pl,
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left;",
      "xG cut7 vs goals-only GLM per league. Log-loss, Brier, RPS improvement (%). Walk-forward CV, 24-month train / 1-month test. All 5 top leagues show improvement. Source: plan_xg_features.R."
    )
  )
} else {
  xg_pl
}
```

```{r vig-dc-vs-glm, echo = FALSE}
dc_xg <- safe_tar_read("vig_dc_vs_glm_xg")
if (is.data.frame(dc_xg) && nrow(dc_xg) > 0) {
  dark_dt(
    dc_xg,
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left;",
      "Three-model comparison on xG-era data (top 5 leagues, 2019+). GLM+xG cut7 improves on both GLM baseline and Dixon-Coles. DC adds the low-score correction (rho) but does not use xG features. Source: plan_xg_features.R."
    )
  )
} else {
  dc_xg
}
```

*xG features make average predictions better calibrated (real, measurable), but this calibration improvement does not translate to profitable per-match AH bet selection. The closing Pinnacle line has already absorbed what xG knows; the model cannot outbid it by enough to clear vig. See [issue #84](https://github.com/JohnGavin/footbet/issues/84) for the full analysis.*

### Market Baselines

```{r vig-market-baselines, echo = FALSE}
mb <- safe_tar_read("vig_market_baselines")
if (is.data.frame(mb) && nrow(mb) > 0) {
  dark_dt(
    mb,
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left;",
      "Pinnacle vs consensus (average bookmaker) calibration per league. Pinnacle is ~1% better on log-loss. Both are the benchmarks all models must beat. Source: plan_evaluation.R."
    )
  )
} else {
  mb
}
```

:::

---

```{r build-info, echo = FALSE, results = "asis"}
safe_tar_read("vig_build_info")
```
